---
title: "annual_data"
author: "Maxine Chapman and Giorgos Papageorgiou"
date: "2025_04_23"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reads in raw data and creates a new output table with various variables for further analysis.

# Setup

```{r}
rm(list =ls()) # clear environment
#get packages
library(MASS)
library(r2r)
library(igraph)
library(raster)
library(scales)
library(sp)
library(sf)
library(adehabitatHR)
#library(rgdal)
library(lubridate) # date and time
library(ggplot2)
library(dplyr)
```

```{r}
source("02_data_processing_functions_MC.R")
```

Read receiver data and set up hashmap for matching receiver ID by lat/lon (to 3 dp) in each shark's data.

```{r}
#get receiver data
receiver_data <- read.csv("../data/raw/acoustic_metadata/darros_receiver_details.csv")
head(receiver_data)

colnames(receiver_data)[which(colnames(receiver_data) == "DSJ.Code")] <- "ReceiverID"

## make lat and lon field with 3 dp 
receiver_data$Lat <- round(receiver_data$Lat, 3)
receiver_data$Long <- round(receiver_data$Long, 3)
receiver_data$Habitat <- as.factor(receiver_data$Habitat) # convert to factor
receiver_data$Location <- as.factor(receiver_data$Location) # convert to factor
str(receiver_data)  # check fields and format
```

```{r}
#Hashmap to identify receiver ID from lat and long
ID_hashmap <- r2r::hashmap() 

for(row in 1:nrow(receiver_data)){
  #row <- 3   #TESTING ONLY
  ID <- receiver_data[row, "ReceiverID"] 
  lat <- receiver_data[row, "Lat"]
  long <- receiver_data[row, "Long"]
  ID_hashmap[[c(lat, long)]] <- ID
}

rm(ID, row, lat, long) # delete temporary objects from the loop to keep environment clean
```

Read in all the individual shark data and combine into one large dataframe.

```{r}
#get all shark data and combine in df
shark_files <- file.path("../data/raw/acoustic_data/", list.files("../data/raw/acoustic_data/"))
df.list <- lapply(shark_files, read.csv, stringsAsFactors=F)
df <- do.call(rbind, df.list)
rm(shark_files, df.list)  # remove unneeded objects

## make new lat and long field with 3 dp and remove the old Lat and Long fields
df$Lat <- round(df$Latitude, 3)
df$Long <- round(df$Longitude, 3)
df <- df[ , -which(colnames(df) %in% c("Latitude", "Longitude"))]

#format shark data as needed, i.e. the datetime column as a datetime object
head(df)
str(df)
df$ISO.date <- as.POSIXct(df$ISO.date, tz = "UTC", format = "%Y-%m-%d %H:%M:%OS")
df$FishID <- as.factor(df$FishID)
str(df)
```

Join shark and receiver data:

```{r}
df <- dplyr::left_join(df, receiver_data)
str(df)
```

Read in summary shark data:

```{r}
summary_sharks <- read.csv("../data/raw/acoustic_metadata/lemon_summary_read.csv")
str(summary_sharks)
head(summary_sharks)

#format dates as needed
summary_sharks$Date.released <- as.POSIXct(summary_sharks$Date.released, tz = "UTC", format = "%d/%m/%Y %H:%M")
summary_sharks$Last.Detection <- as.POSIXct(summary_sharks$Last.Detection, tz = "UTC", format = "%d/%m/%Y %H:%M")
                                           
str(summary_sharks)
head(summary_sharks)
```

# Colour coding for igraphs

Create a list which contains the colour that each node should be plotted on the maps (according to habitat type).

```{r}
#create habitat hashmap
habitat_hashmap <- r2r::hashmap()  # initialise hashmap

for(row in 1:nrow(receiver_data)){
  #row <- 1  #TESTING ONLY
  habitat <- receiver_data[row, "Habitat"]
  lat <- receiver_data[row, "Lat"]
  long <- receiver_data[row, "Long"]
  habitat_hashmap[[c(lat, long)]] <- habitat
}

rm(row, habitat, lat, long)  # clean environment

#initialise list object to store the colour for each receiver
vertex.colours = list()

#run through receiver data
for(receiver in 1:nrow(receiver_data)){
  #receiver <- 1 # TESTING
  
  #get information about receiver
  lat <- receiver_data[receiver, "Lat"] 
  long <- receiver_data[receiver, "Long"]
  habitat_type <- habitat_hashmap[[c(lat, long)]]
  
  #colour coding
  if (habitat_type == "Plateau"){
    vertex.colours <- append(vertex.colours, "orange")
  } 
  else if (habitat_type == "Dropoff"){
     vertex.colours <- append(vertex.colours, "purple")
  }
  else if (habitat_type == "Lagoon") {
      vertex.colours <- append(vertex.colours, "green")
  }
  else if (habitat_type == "Coastal reef"){
      vertex.colours <- append(vertex.colours, "pink")
  }
}

rm(receiver, lat, long, habitat_type)  # clean environment

```


# Igraph subset by year

Carry out functions:

```{r}
data_names_list <- unique(df$FishID)  
years_list <- c(2012:2022)

for (i_shark in 1:length(data_names_list)) {  
  #i_shark <- 15  # TESTING ONLY - leave commented out the rest of the time
  
  #get shark data
  shark_ID <- data_names_list[i_shark]
  shark_file <- df[df$FishID == shark_ID, ]
  
  for(i_year in 1:length(years_list)){
    
    #i_year <- 10 #TESTING ONLY
    
    #subset by year
    this_year <- years_list[i_year]    
    start_year <- paste0(this_year, "-01-01")
    end_year <- paste0(this_year+1, "-01-01")
    
    #subset data
    subset_data <- shark_file[shark_file$ISO.date > start_year & shark_file$ISO.date < end_year, ]
    
    #check there is data for year
    if((nrow(subset_data)) > 1){
      
      #carrying out functions
      shark_matrix <- make_adj_matrix(subset_data, receiver_data, ID_hashmap)
      make_yearly_igraphs(subset_data, shark_matrix, receiver_data, vertex.colours, this_year, shark_ID)
    }
    
  }
  
}
rm(data_names_list, years_list, i_shark, shark_ID, shark_file, i_year, this_year, start_year, end_year, subset_data, shark_matrix)

```

# Results Table

Create list of island v non island receivers:

```{r}
island_rec <- list()
non_island_rec <- list()

#run through reciever data
for(i_receiver in 1:nrow(receiver_data)){
  
  #i_receiver <- 1 #TESTING
  
  #get information about receiver
  lat <- receiver_data[i_receiver, "Lat"]
  long <- receiver_data[i_receiver, "Long"]
  
  #use island long and lat to split receivers into island and non island
  if((53.28 < long)&&(long < 53.38)&&(-5.46 < lat)&&(lat < -5.39)){ #island receivers
    island_rec <- append(island_rec, i_receiver)
  } else{ #non island receivers
    non_island_rec <- append(non_island_rec, i_receiver)
  }
    
}

rm(i_receiver, lat, long)

```

Create list of inner lagoon v outer lagoon receivers:

```{r}
in_lagoon_rec <- list()
out_lagoon_rec <- list()

for(i_receiver in 1:nrow(receiver_data)){
  
  #i_receiver <- 73 #TESTING
  
  #get information about receiver
  lat <- receiver_data[i_receiver, "Lat"] 
  long <- receiver_data[i_receiver, "Long"]
  habitat_type <- habitat_hashmap[[c(lat, long)]]
  
  #sort based on longitude
  if((53.324 < long)&&(long < 53.346)&&(-5.438 < lat)&&(lat < -5.4175)){ #inner lagoon coordinates
    in_lagoon_rec <- append(in_lagoon_rec, i_receiver)
    #print("inner lagoon receiver detected")
  } else if(habitat_type == "Lagoon"){ #island coordinates
    out_lagoon_rec <- append(out_lagoon_rec, i_receiver)
  } 
  
}

rm(i_receiver, lat, long)
```

```{r}
#create empty results dataframe
colnames <- c("shark_ID", 
              "year", 
              "size_slow",
              "size_fast",
              "num_detections", 
              "num_receivers", 
              "n_rec_lagoon", 
              "n_inner_lagoon",
              "n_outer_lagoon",
              "n_rec_plat", 
              "n_rec_drop", 
              "n_rec_coast",
              "min_date", 
              "max_date", 
              "n_tracking_days", 
              "tracking_period",
              "binary_cat",
              "binary_num",
              "sex",
              "sex_num",
              "maturity_slow",
              "mcp_area_100", 
              "mcp_area_95", 
              "mcp_area_80",
              "mcp_area_75",
              "mcp_area_70", 
              "mcp_area_60")
results_df <- data.frame(matrix(nrow = 0, ncol = length(colnames)))
colnames(results_df) <- colnames

head(results_df)
# head(results_df$shark_ID)
```

Make results dataframe:

```{r}
data_names_list <- unique(df$FishID)
years_list <- c(2012:2021)

for (i_shark in 1:length(data_names_list)) {
  
  #i_shark <- 1  # TESTING lizzie
  
  #get shark data
  shark_ID <- data_names_list[i_shark]
  shark_file <- df[df$FishID == shark_ID, ]
  
  #make results dataframe
  for(i_year in 1:length(years_list)){
    #i_year <- 6  # TESTING
    #get dates for 1 year
    this_year <- years_list[i_year]    
    start_year <- paste0(this_year, "-01-01")
    end_year <- paste0(this_year+1, "-01-01")
    
    #subset data
    subset_data <- shark_file[shark_file$ISO.date > start_year & shark_file$ISO.date < end_year, ]
    
    #only run code if there is data available for that year
    if(nrow(subset_data) == 0){
      next #no data skip to next iteration in loop
    }
    
    shark_number <- subset_data$FishID[1]
    year <- this_year
    
    #calculate possible sizes based on different growth rates
    shark_size_slow <- calc_shark_size(shark_number, year, growth_rate = 12.5, summary_sharks)
    shark_size_fast <- calc_shark_size(shark_number, year, growth_rate = 15.5, summary_sharks)
    
    #get number of detections
    num_detections <- nrow(subset_data)
    
    #round data so hashmap can use long and lat to get ID
    subset_data$Lat <- round(subset_data$Lat,3)
    subset_data$Long <- round(subset_data$Long,3)

    #number of unique receivers used
    num_receivers <- nrow(unique(data.frame(id = subset_data$FishID, lon =  subset_data$Long, lat = subset_data$Lat)))
    
    
    #mcp areas
    if(num_receivers > 5){ #calculate if there are more than 5 unique receivers
      mcp_area <- calc_mcp_area(subset_data, 100) #use function to calculate for different percentages
      mcp_area_95 <- calc_mcp_area(subset_data, 95)
      mcp_area_80 <- calc_mcp_area(subset_data, 80)
      mcp_area_75 <- calc_mcp_area(subset_data, 75)
      mcp_area_70 <- calc_mcp_area(subset_data, 70)
      mcp_area_60 <- calc_mcp_area(subset_data, 60)
      
    } else {
      mcp_area <- "NA" #set MCP as NA if there are too few receivers to carry out calculations
      mcp_area_95 <- "NA"
      mcp_area_80 <- "NA"
      mcp_area_75 <- "NA"
      mcp_area_70 <- "NA"
      mcp_area_60 <- "NA"
    }
    
    #dates
    min_date <- min(subset_data$ISO.date)
    max_date <- max(subset_data$ISO.date)
    
    
    tracking_period <- round(difftime(max_date, min_date, units = "days"), 3)
    n_tracking_days <- length(unique(lubridate::date(subset_data$ISO.date)))
    
    #habitats
    shark_unique_recs <- unique(data.frame(ReceiverID = subset_data$ReceiverID, Habitat = subset_data$Habitat))
    
    n_rec_lagoon <- nrow(subset(shark_unique_recs, Habitat == "Lagoon"))
    n_rec_plat <- nrow(subset(shark_unique_recs, Habitat == "Plateau"))
    n_rec_drop <- nrow(subset(shark_unique_recs, Habitat == "Dropoff"))
    n_rec_coast <- nrow(subset(shark_unique_recs, Habitat == "Coastal Reef"))
    
    #binary category - tells us if a shark is an island or non island shark
    
    #get list of receivers used
    receiver_list <- unique(data.frame(id = subset_data$ReceiverID))
    
    for(i_receiver in 1:nrow(receiver_list)){
      #i_receiver <- 14 #TESTING
      #get receiver ID
      ID <- receiver_list[i_receiver, ]
      #check if this is a non island receiver
        if((ID %in% non_island_rec) == TRUE){
          binary_cat <- "exploratory"
          binary_num <- 1
          #stop loop
          break
        } else {
          binary_cat <- "resident"
          binary_num <- 0
        }
    }
    
    in_lagoon <- 0
    out_lagoon <- 0
    
    for(i_receiver in 1:nrow(receiver_list)){
      #get receiver ID
      ID <- receiver_list[i_receiver, ]
      #check location
      if(ID %in% in_lagoon_rec){
        in_lagoon <- in_lagoon + 1
      } else if(ID %in% out_lagoon_rec){
        out_lagoon <- out_lagoon + 1
      }

    }
    
    #get shark sex from summary document
    shark_sex <- summary_sharks[summary_sharks$FishID == shark_ID, "Sex"]
    #shark sex as binary variable 
    if(shark_sex == "m"){
      shark_sex_num <- 0
    } else if((shark_sex == "M")){
      shark_sex_num <- 0
      shark_sex <- "m"
    } else if((shark_sex == "f")){
      shark_sex_num <- 1
    } else if((shark_sex == "F")){
      shark_sex_num <- 1
      shark_sex <- "f"
    }  
    
    #maturity
    if (shark_size_slow > 220){
      maturity_slow <- "mature"
    } else{
      maturity_slow <- "juvenile"
    }
    
    #combine data into row
    new_row <- data.frame("shark_ID" = shark_number,
                          "year" = year, 
                          "size_slow" = shark_size_slow,
                          "size_fast" = shark_size_fast,
                          "num_detections" = num_detections,
                          "num_receivers" = num_receivers, 
                          "n_rec_lagoon" = n_rec_lagoon, 
                          "n_inner_lagoon" = in_lagoon,
                          "n_outer_lagoon" = out_lagoon,
                          "n_rec_plat" = n_rec_plat, 
                          "n_rec_drop" = n_rec_drop, 
                          "n_rec_coast" = n_rec_coast,
                          "min_date" = min_date, 
                          "max_date" = max_date, 
                          "n_tracking_days" = n_tracking_days, 
                          "tracking_period" = tracking_period,
                          "binary_cat" = binary_cat,
                          "binary_num" = binary_num,
                          "sex" = shark_sex,
                          "sex_num" = shark_sex_num,
                          "maturity_slow" = maturity_slow,
                          "mcp_area_100" = mcp_area, 
                          "mcp_area_95" = mcp_area_95, 
                          "mcp_area_80" = mcp_area_80, 
                          "mcp_area_75" = mcp_area_75,
                          "mcp_area_70" = mcp_area_70, 
                          "mcp_area_60" = mcp_area_60)
    
    
    if(length(new_row)!= length(colnames)){
      stop("row is incomplete for shark ", shark_ID,": ", new_row)
    } else {
      
      #add to results df
      results_df <- rbind(results_df, new_row)
      
    }
  }
}

rm(data_names_list, i_year, colnames, start_year, this_year, years_list, end_year) #remove loop variables

#remove variables for results table columns
rm(shark_number, 
   year, 
   shark_size_slow, 
   shark_size_fast,
   num_detections, 
   num_receivers, 
   n_rec_lagoon, 
   in_lagoon,
   out_lagoon,
   n_rec_plat, 
   n_rec_drop, 
   n_rec_coast, 
   min_date, 
   max_date,
   tracking_period,
   n_tracking_days, 
   binary_cat,
   binary_num,
   shark_sex,
   shark_sex_num,
   mcp_area, 
   mcp_area_95, 
   mcp_area_80,
   mcp_area_75,
   mcp_area_70, 
   mcp_area_60)

head(results_df)
str(results_df)

write.csv(results_df, "../results/results_dataframe_annual_23042025.csv", row.names = F)
```

```{r}
print(results_df)

# df_2 <- read.csv("../results/results_dataframe_annual.csv")
# identical(results_df, df_2)
# all.equal(results_df, df_2)
```

# Summary results 

Summary stats
```{r}
# number of sharks
length(unique(results_df$shark_ID))

# sex
sex_counts <- table(unique(results_df[, c("shark_ID", "sex")])$sex)
sex_counts

# total detections
sum(results_df$num_detections, na.rm = TRUE) #total

# number of tracking days (days with at least 1 detection)
tracking_days_per_shark <- aggregate(n_tracking_days ~ shark_ID, data = results_df, sum) # Sum tracking days per shark
mean(tracking_days_per_shark$n_tracking_days)
max(tracking_days_per_shark$n_tracking_days)
min(tracking_days_per_shark$n_tracking_days)


# tracking duration (days from first to last detection)
shark_tracking_summary <- results_df %>%
  group_by(shark_ID) %>%
  summarise(first_detection = min(min_date),
            last_detection = max(max_date),
            total_tracking_period = difftime(last_detection, first_detection, units = "days"))
mean(shark_tracking_summary$total_tracking_period)
min(shark_tracking_summary$total_tracking_period)
max(shark_tracking_summary$total_tracking_period)

# size at release
release_sizes <- results_df %>%
  group_by(shark_ID) %>%
  slice_min(year) %>%  # Get the release year (first year) per shark
  ungroup() %>%
  summarise(
    mean_size = mean(size_slow, na.rm = TRUE),
    max_size = max(size_slow, na.rm = TRUE),
    min_size = min(size_slow, na.rm = TRUE)
  )
release_sizes

# size classes at release
release_stages <- results_df %>%
  group_by(shark_ID) %>%
  slice_min(year) %>%  # Get only release year
  ungroup() %>%
  summarise(
    adults = sum(size_slow >= 220, na.rm = TRUE),
    subadults = sum(size_slow >= 130 & size_slow < 220, na.rm = TRUE),
    juveniles = sum(size_slow <130, na.rm = TRUE)
  )
release_stages

# number of subadults to turn adults
transitions <- results_df %>%
  group_by(shark_ID) %>%
  summarise(
    ever_subadult = any(size_slow < 220 & size_slow > 130, na.rm = TRUE),
    ever_adult = any(size_slow >= 220, na.rm = TRUE)
  ) %>%
  filter(ever_subadult & ever_adult)
transitions$shark_ID

# rm(sex_counts, shark_tracking_summary,release_sizes, release_stages, transitions )
```

Location detections
```{r}
# receiver IDs in non-island receivers
non_island_receiver_ids <- receiver_data$ReceiverID[unlist(non_island_rec)]

# filter detection data
outside_detections <- df[df$ReceiverID %in% non_island_receiver_ids, ]

# non-island receivers per shark
shark_non_island_receivers <- outside_detections %>%
  group_by(FishID) %>%
  summarise(receivers = list(unique(ReceiverID))) %>%
  ungroup()

receiver_lookup <- receiver_data %>% select(ReceiverID, Location) # include receiver location
shark_non_island_receivers_named <- shark_non_island_receivers %>%
  rowwise() %>%
  mutate(location_names = list(receiver_lookup$Location[match(receivers, receiver_lookup$ReceiverID)])) %>%
  ungroup()

# specific locations where sharks were detected 
unique(unlist(shark_non_island_receivers_named$location_names))

# which sharks and when have visited a receiver?
# Choose the receiver you're interested in
target_receiver <- 58 

# Filter the main dataset for that receiver
df_5 <- df %>% 
  filter(ReceiverID == target_receiver) %>%
  mutate(year = year(ISO.date)) %>%
  select(FishID, year) %>%
  distinct() %>%
  arrange(year, FishID)
df_5
```

General detections figure
```{r}
# str(df)
detection_counts <- df %>%
  group_by(FishID) %>%
  summarise(n = n())

ggplot(df, aes(x = ISO.date, y = FishID, colour = Habitat)) +
  geom_point(alpha = 1, size = 0.2) +
  geom_text(
    data = detection_counts,
    aes(x = max(df$ISO.date)+ 90, y = FishID, label = n), 
    inherit.aes = FALSE,
    size = 2,
    hjust = 0
  )+
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Date",
    y = "Shark ID",
    colour = "Habitat"
  ) +
  xlim(min(df$ISO.date), max(df$ISO.date) + 150) +
  theme_minimal(base_size=14)+
  theme(
    axis.text.y = element_text(size=6),
    legend.position = "bottom"
  )

```

```{r}
ggsave("../results/all_detections.pdf", width = 10, height = 6)
```


