---
title: "COAs"
author: "Giorgos Papageorgiou"
date: "2026-08-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up and data prep
```{r}
rm(list =ls())
library(tidyverse)
library(lubridate)
library(sf)
```

## Read detections
```{r}
df <- read.csv("../data/intermediate/combined_sharks.csv")
```

## Timestamps
Observations without a complete timestamp are excluded because COA calculation requires both date and time.
```{r}
df <- df %>%
  mutate(
    datetime = ymd_hms(ISO.date, tz = "UTC", quiet = TRUE),
    year = year(datetime))

df <- df %>%
  filter(!is.na(datetime),
         year >= 2012,
         year <= 2021)
```

## Receiver locations and projections
```{r}
# create one row per unique receiver location
receiver_locs <- df %>%
  distinct(ReceiverID, Long, Lat)

# number of unique receiver locations
nrow(receiver_locs)

# check whether any ReceiverID has more than one coordinate pair
receiver_check <- receiver_locs %>%
  count(ReceiverID) %>%
  filter(n > 1)
stopifnot(nrow(receiver_check) == 0)
```

Receiver coordinates are projected from WGS84 longitude/latitude into a local LAEA coordinate system. 
This allows COAs to be calculated from projected X and Y coordinates in metres, rather than directly averaging geographic coordinates in degrees. 
```{r}
# convert receiver locations to sf object
receiver_sf <- st_as_sf(
  receiver_locs,
  coords = c("Long", "Lat"),
  crs = 4326,
  remove = FALSE)

# project to the same local LAEA CRS used for the MCP analysis
laea_crs <- "+proj=laea +lat_0=-5.4 +lon_0=53 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"

receiver_sf <- st_transform(receiver_sf, crs = laea_crs)

# extract projected X and Y coordinates
xy <- st_coordinates(receiver_sf)

receiver_sf <- receiver_sf %>%
  mutate(
    X = xy[, 1],
    Y = xy[, 2])
```

Projected receiver coordinates joined back to the main dataset
```{r}
# extract projected coordinates without sf geometry
receiver_xy <- receiver_sf %>%
  st_drop_geometry() %>%
  select(ReceiverID, X, Y)

# join projected coordinates onto every detection
df <- df %>%
  left_join(receiver_xy, by = "ReceiverID")

# confirm that all detections received projected coordinates
stopifnot(
  all(!is.na(df$X)),
  all(!is.na(df$Y)))
```

# Choose COA time interval
Test 30, 60, 90 and 120 min before calculating the final COAs
```{r}
intervals <- c(30, 60, 90, 120)

interval_summary <- lapply(intervals, function(int) {
  
  tmp <- df %>%
    mutate(
      test_bin = as.POSIXct(
  floor(as.numeric(datetime) / (int * 60)) * (int * 60),
  origin = "1970-01-01",
  tz = "UTC")) %>%
    group_by(FishID, test_bin) %>%
    summarise(
      n_receivers = n_distinct(ReceiverID),
      .groups = "drop")
  
  data.frame(
    interval_min = int,
    n_bins = nrow(tmp),
    percent_1_receiver = mean(tmp$n_receivers == 1) * 100,
    percent_2plus = mean(tmp$n_receivers >= 2) * 100,
    percent_3plus = mean(tmp$n_receivers >= 3) * 100)
  }) %>%
  bind_rows()

interval_summary
```

## Detection frequency
```{r}
interval_detail <- lapply(intervals, function(int) {
  
  tmp <- df %>%
    mutate(
      test_bin = as.POSIXct(
        floor(as.numeric(datetime) / (int * 60)) * (int * 60),
        origin = "1970-01-01",
        tz = "UTC"
      )
    ) %>%
    group_by(FishID, test_bin) %>%
    summarise(
      n_receivers = n_distinct(ReceiverID),
      n_detections = n(),
      .groups = "drop"
    )
  
  data.frame(
    interval_min = int,
    median_detections = median(tmp$n_detections),
    mean_detections = mean(tmp$n_detections),
    median_receivers = median(tmp$n_receivers),
    percent_2plus = mean(tmp$n_receivers >= 2) * 100
  )
  
}) %>%
  bind_rows()

interval_detail
```

## Receiver separation
```{r}
calc_bin_span <- function(data, interval_min) {
  
  tmp <- data %>%
    mutate(
      test_bin = as.POSIXct(
        floor(as.numeric(datetime) / (interval_min * 60)) *
          (interval_min * 60),
        origin = "1970-01-01",
        tz = "UTC"))
  
  tmp %>%
    group_by(FishID, test_bin) %>%
    filter(n_distinct(ReceiverID) >= 2) %>%
    summarise(
      n_receivers = n_distinct(ReceiverID),
      max_receiver_distance_m = {
        coords <- cbind(X, Y)
        max(as.matrix(dist(unique(coords))))}, .groups = "drop") %>%
    mutate(interval_min = interval_min)
}
```

```{r}
span_all <- lapply(c(30, 60, 90, 120), function(int) {
  calc_bin_span(df, int)
}) %>%
  bind_rows()

span_summary <- span_all %>%
  group_by(interval_min) %>%
  summarise(
    n_multi_receiver_bins = n(),
    median_distance_m = median(max_receiver_distance_m),
    mean_distance_m = mean(max_receiver_distance_m),
    p95_distance_m = quantile(max_receiver_distance_m, 0.95),
    max_distance_m = max(max_receiver_distance_m),
    .groups = "drop")

span_summary
```

# Calculate final 60-minute COAs

Based on the interval comparison above, a 60-minute interval is selected for the final COA calculation. For each shark and 60-minute interval, detections are counted at each receiver. Receiver coordinates are then weighted by the number of detections at that receiver to calculate the mean projected X and Y position. For Intervals containing detections at only one receiver, the COA corresponds to the location of that receiver.

## Calculation
```{r}
# final COA interval
coa_interval <- 60

# assign detections to 60-minute bins and count detections per receiver
coa_counts_final <- df %>%
  mutate(
    coa_bin = as.POSIXct(
      floor(as.numeric(datetime) / (coa_interval * 60)) *
        (coa_interval * 60),
      origin = "1970-01-01",
      tz = "UTC")) %>%
  group_by(FishID, coa_bin, ReceiverID, X, Y) %>%
  summarise(
    n_detections = n(),.groups = "drop")

coa <- coa_counts_final %>%
  group_by(FishID, coa_bin) %>%
  summarise(
    coa_X = weighted.mean(X, n_detections),
    coa_Y = weighted.mean(Y, n_detections),
    n_receivers = n_distinct(ReceiverID),
    n_detections = sum(n_detections),
    .groups = "drop")
```

## Transform to long/lat 

COAs were transformed back to WGS84 longitude and latitude so that the resulting pseudo-positions could be passed to the tidytracks KDE workflow
```{r}
# convert COA coordinates to an sf object in the projected CRS
coa_sf <- st_as_sf(
  coa,
  coords = c("coa_X", "coa_Y"),
  crs = laea_crs,
  remove = FALSE)

# transform back to WGS84 longitude/latitude
coa_sf_wgs84 <- st_transform(coa_sf, crs = 4326)

# extract longitude and latitude
coa_lonlat <- st_coordinates(coa_sf_wgs84)

coa <- coa %>%
  mutate(
    Long = coa_lonlat[, 1], Lat  = coa_lonlat[, 2])
```

```{r}
coa_kde <- coa %>%
  transmute(
    FishID,
    ISO.date = format(coa_bin, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"), 
    Long, Lat, coa_X, coa_Y, n_receivers, n_detections)
```

## Validate dataset
```{r}
# Every original detection should contribute to exactly one COA
stopifnot(sum(coa$n_detections) == nrow(df))

# Every COA should contain at least one detection and one receiver
stopifnot(all(coa$n_detections >= 1))
stopifnot(all(coa$n_receivers >= 1))

# Number of receivers cannot exceed number of detections
stopifnot(all(coa$n_receivers <= coa$n_detections))

# No missing COA coordinates
stopifnot(
  all(!is.na(coa$coa_X)),
  all(!is.na(coa$coa_Y)),
  all(!is.na(coa$Long)),
  all(!is.na(coa$Lat)))

# One COA only per shark per time bin
stopifnot(
  coa %>%
    count(FishID, coa_bin) %>%
    filter(n > 1) %>%
    nrow() == 0)

# All sharks in the detection dataset are retained
stopifnot(
  setequal(unique(df$FishID), unique(coa$FishID)))

# select one multi-receiver COA for a manual check
test_bin <- coa %>%
  dplyr::filter(n_receivers >= 2) %>%
  dplyr::slice(1)

test_counts <- coa_counts_final %>%
  filter(
    FishID == test_bin$FishID,
    coa_bin == test_bin$coa_bin)

# manually reproduce weighted mean
manual_X <- weighted.mean(test_counts$X, test_counts$n_detections)
manual_Y <- weighted.mean(test_counts$Y, test_counts$n_detections)

stopifnot(
  isTRUE(all.equal(manual_X, test_bin$coa_X)),
  isTRUE(all.equal(manual_Y, test_bin$coa_Y)))
```

## Export
```{r}
write.csv(
  coa_kde,
  "../data/intermediate/combined_sharks_coa.csv",
  row.names = FALSE)
```
