---
title: "Annual space-use metrics"
author: "Maxine Chapman"
date: "2026-07-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Creating annual results df

This code creates an annual dataframe with one row per shark per year. Each row contains basic tracking information on the shark and calculates area metrics including kde and mcp. 

## Setup
```{r cars}
rm(list =ls())
library(tidytracks)
library(dplyr)
library(lubridate)
```

```{r}
# raw detection data
df <- read.csv("../data/intermediate/combined_sharks.csv")

# 60-minute centre-of-activity positions
coa_df <- read.csv("../data/intermediate/combined_sharks_coa.csv")

# local equal-area projection used for spatial area calculations
local_laea <- "+proj=laea +lat_0=-5.4 +lon_0=53 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"

study_years <- 2012:2021
maturity_cutoff <- 166
```

```{r}
df$year <- lubridate::year(as.POSIXct(df$ISO.date, tz = "UTC", format = "%Y-%m-%d %H:%M:%OS"))
coa_df$year <- lubridate::year(as.POSIXct(coa_df$ISO.date, tz = "UTC", format = "%Y-%m-%d %H:%M:%S"))

# Remove detections without a complete date-time
df <- df %>%
  filter(!is.na(year))

stopifnot(!any(is.na(df$year)))
stopifnot(!any(is.na(coa_df$year)))
```

```{r}
#add in shark metadata
summary_sharks <- read.csv("../data/raw/acoustic_metadata/lemon_summary_read_updated.csv")
```

Read in our data processing functions which include calculating MCP area and algorithms to estimate shark size in a given year, using a VBGF curve with precaudal length based on Brown and Gruber (1988).

```{r}
source("03_data_processing_functions_MC.R")
```

## Create results df

Create the annual results dataframe. This stores one row per shark per year, including tracking effort, estimated size, sex, maturity, and annual space-use metrics.
```{r}
cols <- c("shark_ID", 
              "year", 
              "size",
              "num_detections", 
              "num_receivers", 
              "min_date", 
              "max_date", 
              "n_tracking_days",
              "n_detection_months",
              "max_internal_gap_days",
              "tracking_period",
              "sex",
              "maturity",
              "n_kde_positions",
              "kde_area_95",
              "kde_area_50",
              "mcp_area_100")
results_df <- data.frame(matrix(nrow = 0, ncol = length(cols)))
colnames(results_df) <- cols
```

```{r}
data_names_list <- unique(df$FishID)
years_list <- study_years
```

Create dataframe for maximum gap between detection days (annually). This will be joined to the metadata table in the metadata script
```{r}
# one row per shark: maximum gap between detection days
detection_gap_df <- data.frame(
  FishID = character(),
  max_detection_gap_days = integer())
```

```{r}
for (i_shark in 1:length(data_names_list)) {
  
  #get shark data
  shark_ID <- data_names_list[i_shark]
  shark_file <- df[df$FishID == shark_ID, ]
  # get COA data for the same shark
  shark_coa_file <- coa_df[coa_df$FishID == shark_ID, ]
  
  # calculate maximum gap between consecutive detection days
  detection_days <- shark_file %>%
    mutate(detection_date = as.Date(ISO.date)) %>%
    distinct(detection_date) %>%
    arrange(detection_date)

    max_gap <- if (nrow(detection_days) > 1) {
      max(as.integer(diff(detection_days$detection_date)) - 1) } else { NA_integer_}

    detection_gap_df <- rbind(
      detection_gap_df,
      data.frame(FishID = shark_ID, max_detection_gap_days = max_gap))
  
  #make results dataframe
  for(i_year in 1:length(years_list)){
    
    #get dates for 1 year
    this_year <- years_list[i_year]    
    start_year <- paste0(this_year, "-01-01")
    end_year <- paste0(this_year+1, "-01-01")
    
    #subset data
    subset_data <- shark_file[shark_file$ISO.date > start_year & shark_file$ISO.date < end_year, ]
    # subset COA positions for the same shark and year
    subset_coa <- shark_coa_file[shark_coa_file$ISO.date > start_year &shark_coa_file$ISO.date < end_year, ]
    
    # only run code if there is data available for that year
    if(nrow(subset_data) == 0){
      next # no data skip to next iteration in loop
    }
    
    shark_number <- subset_data$FishID[1]
    year <- this_year
    
    #calculate shark size using function
    shark_size <- calc_shark_size(shark_number, year, summary_sharks)
    
    #get number of detections
    num_detections <- nrow(subset_data)
    
    #round data so hashmap can use long and lat to get ID
    subset_data$Lat <- round(subset_data$Lat,3)
    subset_data$Long <- round(subset_data$Long,3)

    #number of unique receivers used
    num_receivers <- nrow(unique(data.frame(id = subset_data$FishID, lon =  subset_data$Long, lat = subset_data$Lat)))
    
    # use 60-minute COA positions for spatial analyses
    subset_kde <- subset_coa

    if (nrow(subset_kde) == 0) {
      stop("No COA positions available for shark ",shark_number, " in ", year)}

    kde_method <- "COA_60min"
    n_kde_positions <- nrow(subset_kde)
    
    #kde areas
    subset_tt <- tt_read_data(
      events = subset_kde,
      col_track_id = "FishID",
      col_coords = c("Long", "Lat"),
      col_date_time = "ISO.date",
      time_zone = "UTC",
      format_date_time = "%Y-%m-%d %H:%M:%OS",
      crs = 4326)
    
    # project COA positions to local LAEA before KDE
    # project before KDE
    subset_tt <- sf::st_transform(subset_tt, crs = local_laea)
    
    coords_range <- diff(range(sf::st_coordinates(subset_tt)[, 1])) +
      diff(range(sf::st_coordinates(subset_tt)[, 2]))
    
    if (coords_range == 0) {
      kde_area_50 <- NA_real_
      kde_area_95 <- NA_real_
    } else {
      kde_data <- subset_tt %>%
        hr_kde(levels = c(0.5, 0.95)) %>%
        mutate(area = set_units(area, "km^2"))
      
      kde_area_50 <- kde_data$area[kde_data$level == 0.5]
      kde_area_95 <- kde_data$area[kde_data$level == 0.95]
      kde_area_50 <- if (length(kde_area_50) == 0) NA_real_ else as.numeric(kde_area_50)
      kde_area_95 <- if (length(kde_area_95) == 0) NA_real_ else as.numeric(kde_area_95)
    }

    # MCP calculations from the same 60-min COAs used for KDE
    n_unique_coa_positions <- nrow(unique(subset_coa[, c("Long", "Lat")]))

    if (nrow(subset_coa) >= 5 && n_unique_coa_positions >= 3) {
        mcp_area <- calc_mcp_area(subset_coa, 100)
        mcp_area_95 <- calc_mcp_area(subset_coa, 95)
        mcp_area_90 <- calc_mcp_area(subset_coa, 90)
        mcp_area_80 <- calc_mcp_area(subset_coa, 80)
        mcp_area_75 <- calc_mcp_area(subset_coa, 75)
        mcp_area_70 <- calc_mcp_area(subset_coa, 70)
        mcp_area_60 <- calc_mcp_area(subset_coa, 60)
    } else {
        mcp_area <- NA_real_
      mcp_area_95 <- NA_real_
      mcp_area_90 <- NA_real_
      mcp_area_80 <- NA_real_
      mcp_area_75 <- NA_real_
      mcp_area_70 <- NA_real_
      mcp_area_60 <- NA_real_
}
    
    #dates
    min_date <- min(subset_data$ISO.date)
    max_date <- max(subset_data$ISO.date)
    
    tracking_period <- round(difftime(max_date, min_date, units = "days"), 3)
    
    # unique calendar days on which shark was detected
    detection_dates <- sort(unique(lubridate::date(subset_data$ISO.date)))
    # number of detection days
    n_tracking_days <- length(detection_dates)
    # number of calendar months containing at least one detection
    n_detection_months <- length(unique(lubridate::month(detection_dates)))
    # longest period without a detection between two detected days
    if (length(detection_dates) > 1) {
      max_internal_gap_days <- max(as.numeric(diff(detection_dates)) - 1)
    } else {
    max_internal_gap_days <- NA_real_}

    #get shark sex from summary document
    shark_sex <- summary_sharks[summary_sharks$FishID == shark_ID, "Sex"]

    if (shark_size >= maturity_cutoff){
          maturity <- "mature"
    } else {
          maturity <- "immature"
    }
    
    #combine data into row
    new_row <- data.frame("shark_ID" = shark_number,
                          "year" = year, 
                          "size" = shark_size,
                          "num_detections" = num_detections,
                          "num_receivers" = num_receivers, 
                          "min_date" = min_date, 
                          "max_date" = max_date, 
                          "n_tracking_days" = n_tracking_days,
                          "n_detection_months" = n_detection_months,
                          "max_internal_gap_days" = max_internal_gap_days,
                          "tracking_period" = tracking_period,
                          "sex" = shark_sex,
                          "maturity" = maturity,
                          "n_kde_positions" = n_kde_positions,
                          "kde_area_95" = kde_area_95,
                          "kde_area_50" = kde_area_50,
                          "mcp_area_100" = mcp_area)
    
    
    if(length(new_row)!= length(cols)){
      stop("row is incomplete for shark ", shark_ID,": ", new_row)
    } else {
      
      #add to results df
      results_df <- rbind(results_df, new_row)
      
    }
  }
}
```

Save results df
```{r}
write.csv(
  results_df,
  "../results/results_annual_kde_coa.csv",
  row.names = FALSE
)
```

Save max no detection gap
```{r}
write.csv(
  detection_gap_df,
  "../results/metadata/shark_detection_gaps.csv",
  row.names = FALSE
)
```
