---
title: "2023_01_25_plotting_MC"
author: "Maxine Chapman and Giorgos Papageorgiou"
date: "2023-01-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Plots based on the annual results dataframe (not raw data directly).

# Setup

```{r}
rm(list =ls()) #clear environments
```


## packages

```{r}
library(ggplot2)
library(ggsignif)
library(lme4)
library(performance)
library(cowplot)
library(dplyr)
library(ggsci)
library(geomtextpath)
library(forcats)
library(r2r)
library(magick)
```

## read in data

```{r}
#cutoff data with less than 40 tracking days
results_df <- read.csv("../results/results_dataframe_annual.csv")
plot_df <- results_df[results_df$n_tracking_days > 40, ]
```

```{r}
#total tracking days per shark
summary_df <- plot_df %>% group_by(shark_ID) %>% summarize(total_tracking_period = sum(tracking_period, na.rm = TRUE))
```

```{r}
#read in receiver data
receiver_data = read.csv("../data/raw/acoustic_metadata/darros_receiver_details.csv")
#round lat and lon to 3 sf for matching with the shark data
receiver_data$Lat<-round(receiver_data$Lat,3)
receiver_data$Long<-round(receiver_data$Long,3)

#rename receiver ID column for clarity
colnames(receiver_data)[which(colnames(receiver_data) == "DSJ.Code")] <- "ReceiverID"

#head(receiver_data)
```


## set key variables

```{r}
percent <- 80 #set percentage value of mcp to use for analyses
raw_percent_data <- plot_df$mcp_area_80  #select corresponding mcp data for plotting
percent_data <- raw_percent_data + 1

growth_rate <- "slow" #12.5cm per year growth rate is slow/15.5cm per year is fast
size <- plot_df$size_slow 

#make filepath
filepath <- file.path("..", "results", "mcp_plots", percent)
filepath1 <- file.path("..", "results", "mcp_plots", "80", "mcp_80")
filepath2 <-file.path("..", "results", "figures")

#values without cutoff data(for combined plot)
percent_data_2 <- (results_df$mcp_area_80) + 1 
tracking <- results_df$n_tracking_days
binary <- results_df$binary_cat
size_2 <- results_df$size_slow

#create new file for MCP % if it doesn't exist yet, to save plots in
if (! dir.exists(file.path("..", "results", "mcp_plots", percent))) {
  dir.create(file.path("..", "results", "mcp_plots", percent))
}
```


# Main Figures

## fig. 1: Receiver maps

```{r}
#read in figures
img1 <- magick::image_read("../results/habitat_receiver_plot_big.png")
img2 <- magick::image_read("../results/habitat_receiver_plot.png")
img3 <- magick::image_read("../results/binary_receiver_plot_big.png")
img4 <- magick::image_read("../results/binary_receiver_plot.png")

#convert to ggplot
g1 <- cowplot::ggdraw() + cowplot::draw_image(img1) + theme(plot.margin = margin(5, 5, 5, 0))
g2 <- cowplot::ggdraw() + cowplot::draw_image(img2) + theme(plot.margin = margin(5, 0, 5, 5)) 
g3 <- cowplot::ggdraw() + cowplot::draw_image(img3) + theme(plot.margin = margin(5, 5, 5, 0))
g4 <- cowplot::ggdraw() + cowplot::draw_image(img4) + theme(plot.margin = margin(5, 0, 5, 5)) 
```

```{r}
filename <- "figure_1_receiver_maps.png"

cowplot::plot_grid(g1, g2, g3, g4,
                   labels = "auto",   
                   nrow = 2 
                   #rel_widths = c(1, 1.1)
                   )

ggsave(filename, path = filepath2)
rm(filename)
```

## fig. 2: Descriptive Plot

```{r}
#make new columns
included_sharks <- unique(plot_df$shark_ID)

#plot_df <- plot_df %>% #over 40 tracking days
describe_df <- results_df %>% #includes all sharks
  dplyr::mutate(log_detections = log10(num_detections)) %>%
  dplyr::group_by(shark_ID) %>%
  dplyr::mutate(start_year = min(year)) %>%
  dplyr::mutate(end_year = max(year)) %>%
  dplyr::mutate(total_days = sum(n_tracking_days)) %>%
  dplyr::mutate(included = ifelse(shark_ID %in% included_sharks, "included", "not included")) %>% 
  arrange(desc(included), shark_ID)
  #dplyr::mutate(shark_ID = factor(shark_ID, levels = unique(shark_ID)))

describe_df$shark_ID <- factor(describe_df$shark_ID, levels = unique(describe_df$shark_ID))

maturity_df <- describe_df %>%
  filter(maturity_slow == "mature") %>%
  group_by(shark_ID) %>%
  summarize(maturity_year = min(year), .groups = "drop")

# Step 3: Merge maturity year back into main dataframe
describe_df <- left_join(describe_df, maturity_df, by = "shark_ID")


shade_df <- describe_df %>%
  filter(included == "not included") %>%
  distinct(shark_ID) %>%
  mutate(
    xmin = as.numeric(shark_ID) - 0.5,
    xmax = as.numeric(shark_ID) + 0.5,
    ymin = -Inf,   # cover all years
    ymax = Inf
  )
```

NB: The plot appears differently (less cramped) in saved files than in R due to different dimensions in save.

```{r}
ggplot(describe_df, aes(x = factor(shark_ID), y = year, ymin = year, ymax = year + 1, fill = log_detections)) +
  geom_crossbar(aes(size = NA), width = 0.6) +  # narrower bars
  geom_text(label = "Tracking days", y = 2024, x = length(unique(describe_df$shark_ID)) + 1, hjust = 0.6) + #label tracking days column
  coord_flip() +
  scale_fill_gradient(low = "#999999", high = "#377EB8") +
  scale_y_continuous(breaks = 2012:2022, expand = expansion(mult = c(0, 0.15))) +
  #scale_y_discrete(expand = expansion(add = c(0, 2))) +
  scale_x_discrete(expand = expansion(add = c(0, 2))) +  #more horizontal space for shark_ID
  #labs(title = "Shark Tracking Data Visualisation", x = "Shark ID", y = "Year", fill = "log(number of detections)") +
  labs(x = "Shark ID", y = "Year", fill = "log (number of detections)") +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(size = FALSE) +
  geom_text(aes(y = 2024, label = total_days), vjust = 0.5) +
  theme(panel.grid.major.y = element_line(linewidth = 1, colour = "grey"), axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_bw() +
  geom_rect(data = shade_df,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey80", alpha = 0.5,
            inherit.aes = FALSE) + #grey out irrelevant data
  geom_segment(
  data = maturity_df,
  aes(
    x = shark_ID,
    xend = shark_ID,
    y = maturity_year - 0.05,
    yend = maturity_year + 0.05
  ),
  color = "#F8766D",
  size = 1,
  inherit.aes = FALSE
)
  
#saving
#filename <- "figure_2_descriptive.png"
filename <- "figure_6_descriptive_fast.png"
ggsave(filename, path = filepath2, height = 12)
```

## fig. 3: tracking days v mcp plot

Justify choice to cutoff number of tracking days > 40.

```{r}
#combine plots with and without cutoff for number of tracking days

#create filename for saving
filename <- "figure_7_tracking_days_fast.png"
#filename <- "figure_3_tracking_days.png"

png(paste0(filepath2, "/", filename), width = 3000, height = 2000, res = 300)

#pdf(paste0(filepath, "/", filename))

#plot without cutoff
days <- ggplot2::ggplot(data =  results_df) +
  ggplot2::geom_point(aes(x = tracking, y = percent_data_2)) +
  #ggplot2::ggtitle(paste0("Tracking days(all) v ", percent, "% MCP")) +
  geom_smooth(aes(x = tracking, y = percent_data_2), se = FALSE, span = 0.8, colour = "#377EB8") +
  ylab(paste0("Log (", percent, "% MCP area)")) +
  xlab("Tracking days") +
  #labs(col = "Shark ID") + 
  scale_y_log10(limits = c(1, 150)) +
  scale_x_continuous(limits = c(0, NA)) +
  #guides(color = FALSE) + # Hide legend for color
  geom_vline(xintercept = 40, colour  = "#F8766D", linetype = "longdash", size = 1.2) + #add 40 day line
  annotate("text", x = 40, y = 1, label = "40", colour = "#F8766D", vjust = 4, size = 3.5) +
  coord_cartesian(clip = 'off') +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"), 
        plot.margin = margin(10, 10, 30, 10)) 

#plot with 40 day cutoff
days_limited <- ggplot2::ggplot(data =  plot_df) +
  ggplot2::geom_point(aes(x = n_tracking_days, y = percent_data)) +
  #ggplot2::ggtitle(paste0("Tracking days(>40) v ", percent, "% MCP")) +
  ylab(paste0("Log (", percent, "% MCP area)")) +
  xlab("Tracking days") +
  #labs(col = "Shark ID") + 
  geom_smooth(aes(x = n_tracking_days, y = percent_data), se = FALSE, span = 0.8, colour = "#377EB8") +
  scale_y_log10(limits = c(1, 150)) +
  #scale_x_continuous(limits = c(0, NA)) +
  #guides(color = FALSE) + # Hide legend for color
  # geom_vline(xintercept = 40, colour  = "red", linetype = "longdash", size = 1.2) +
  # annotate("text", x = 40, y = 1, label = "40", colour = "red", vjust = 3, size = 3.5) +
  # coord_cartesian(clip = 'off') +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"), 
        plot.margin = margin(10, 10, 30, 10)) 

#display plots
combined_legend <- cowplot::get_legend(days_limited)
cowplot::plot_grid(days, days_limited, labels = "auto", align = "h", axis = "tb")

#stop save
dev.off()

#remove uneccessary objects
rm(filename)
```

## fig. 4: shark size v mcp area

```{r}
#plotting
ggplot2::ggplot(data = plot_df) +
  #ggplot2::geom_point(aes(x = size, y = percent_data, colour = factor(shark_ID))) +
  ggplot2::geom_point(aes(x = size_slow, y = percent_data)) +
  #ggplot2::ggtitle(paste0("Shark size v ", percent, "% MCP area")) +
  scale_y_log10() +
  ylab(paste0("Log (", percent,"% MCP area)")) +
  xlab("Estimated shark size (cm)") +
  #theme(text = element_text(size = 16)) +
  #geom_abline(intercept =  -0.2513395, slope = 0.003383697, colour = "red") 
  geom_abline(intercept = -0.50948, slope = 0.0046877, colour = "#F8766D") + #mlm result
  # geom_line(data = pred_df, aes(x = size_slow, y = 10^predicted - 1), colour = "#F8766D", size = 1) +
  theme_bw()


#creating filepaths and name
filename <- "figure_4_size_mcp.png"

ggsave(filename, path = filepath2)

rm(filename)
```

## fig. 5: boxplots for resident v exploratory

```{r}
#counts for sample sizes
count <- plot_df %>%
  group_by(binary_cat) %>%
  summarise(count = n()) 

count <- count %>%
  mutate(label = paste(binary_cat, "\nn =", count))
```

```{r}
ggplot2::ggplot(data = plot_df, aes(x = binary_cat, y = size, fill = binary_cat)) +
  ggplot2::geom_boxplot() +
  #2ggplot2::ggtitle("Shark Size v Behaviour") +
  xlab("Behaviour") +
  ylab(paste0("Estimated size (", growth_rate, " growth rate)")) +
  #theme(text = element_text(size = 16)) +
  geom_hline(yintercept = 220, linetype = "dashed", colour = "#F8766D") +
  annotate("text", x = 2.5, y = 220, label = "size at maturity",
           color = "#F8766D", hjust = 0, vjust = -0.5) +
  #scale_x_discrete(expand = expansion(add = c(0.8, 1.2))) +
  scale_x_discrete(labels = setNames(count$label, count$binary_cat), expand = expansion(add = c(0.8, 1.2))) +
  #scale_fill_npg() +
  scale_fill_manual(values = c("#F8766D", "#377EB8")) +
  theme_bw() +
  theme(legend.position = "none")


#filename <- "figure_5_resident_exploratory.png"
filename <- "figure_9_resident_exploratory_fast.png"


ggsave(filename, path = filepath2)
rm(filename)

```

# Supplementary Figures

## fig 6. shark_size v mcp area with FAST growth rate

```{r}
size_fast <- plot_df$size_fast
```


```{r}
#plotting
ggplot2::ggplot(data = plot_df) +
  #ggplot2::geom_point(aes(x = size, y = percent_data, colour = factor(shark_ID))) +
  ggplot2::geom_point(aes(x = size_fast, y = percent_data)) +
  #ggplot2::ggtitle(paste0("Shark size v ", percent, "% MCP area")) +
  scale_y_log10() +
  ylab(paste0("Log (", percent,"% MCP area)")) +
  xlab("Estimated shark size (cm)") +
  #theme(text = element_text(size = 16)) +
  geom_abline(intercept = -0.397855, slope = 0.003951, colour = "#F8766D") +
  theme_bw()


#creating filepaths and name
filename <- "figure_8_size_mcp.png"

ggsave(filename, path = filepath2)

rm(filename)
```

## fig. 10. maturity v range size for male/female

```{r}
#female shark data
female_df <- plot_df[plot_df$sex == "f", ]

#male shark data
male_df <- plot_df[plot_df$sex == "m", ]
```

```{r}
#create filename for saving
filename <- "figure_10_sex.png"

png(paste0(filepath2, "/", filename), width = 2000, height = 2400, res = 300)
#pdf(paste0(filepath, "/", filename))

#female_plot
female <- ggplot2::ggplot(data = female_df, aes(x = maturity_slow, y = log(mcp_area_80), fill = maturity_slow)) +
  ggplot2::geom_boxplot() +
  xlab("Female maturity (>220cm)") +
  ylab((paste0("Log (80% MCP area)"))) +
  scale_fill_manual(values = c("#F8766D", "#377EB8")) +
  scale_y_continuous(limits = c(-2.5, 3)) +
  theme_bw() +
  theme(legend.position = "none")

#male_plot
male <- ggplot2::ggplot(data = male_df, aes(x = maturity_slow, y = log(mcp_area_80), fill = maturity_slow)) +
  ggplot2::geom_boxplot() +
  xlab("Male maturity (>220cm)") +
  ylab((paste0("Log (80% MCP area)"))) +
  scale_fill_manual(values = c("#F8766D", "#377EB8")) +
  scale_y_continuous(limits = c(-2.5, 3)) +
  theme_bw() +
  theme(legend.position = "none")

#display plots
cowplot::plot_grid(female, male, labels = "auto", align = "h", axis = "tb")

#stop save
#dev.off()

#remove uneccessary objects
rm(filename)
```

## Table 1. Average MCP by size class

```{r}
table_1 <- plot_df %>%
  mutate(
    maturity_class = factor(case_when(
      size_slow < 130 ~ "juvenile",
      size_slow >= 130 & size_slow <= 220 ~ "subadult",
      size_slow > 220 ~ "adult",
      TRUE ~ NA_character_
    ), levels = c("juvenile", "subadult", "adult"))
  ) %>%
  group_by(maturity_class) %>%
  summarise(
    size = dplyr::first(c("< 130", "130–220", "> 220")[match(maturity_class, c("juvenile","subadult","adult"))]),
    # MCP_100 = signif(mean(mcp_area_100, na.rm = TRUE), 2), #2sf 
    # MCP_95 = signif(mean(mcp_area_95, na.rm = TRUE), 2),
    # MCP_80 = signif(mean(mcp_area_80, na.rm = TRUE), 2),
    # IQR_80 = signif(IQR(mcp_area_80, na.rm = TRUE), 2),
    # MCP_70 = signif(mean(mcp_area_70, na.rm = TRUE), 2),
    # MCP_60 = signif(mean(mcp_area_60, na.rm = TRUE), 2)
    MCP_100 = round(mean(mcp_area_100, na.rm = TRUE), 1), #1dp
    MCP_90  = round(mean(mcp_area_90, na.rm = TRUE), 1),
    MCP_80  = round(mean(mcp_area_80, na.rm = TRUE), 1),
    IQR_80  = round(IQR(mcp_area_80, na.rm = TRUE), 1),
    MCP_70  = round(mean(mcp_area_70, na.rm = TRUE), 1),
    MCP_60  = round(mean(mcp_area_60, na.rm = TRUE), 1),
  )

write.csv(table_1, "../results/table_1.csv", row.names = F)
```


