---
title: "Annual space-use filtering"
author: "Giorgos Papageorgiou"
date: "2026-09-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
```{r}
rm(list =ls())
library(tidyverse)
```

# Filtering

The annual KDE modelling dataset is filtered using explicit technical and temporal criteria before any models are fitted. The same criteria are applied to all shark-years
```{r}
# Main annual dataset: COA-based KDEs
complete_df <- read.csv("../results/results_annual_kde_coa.csv",
                        stringsAsFactors = FALSE)

complete_df <- complete_df %>%
  mutate(
    min_date = as.Date(min_date),
    max_date = as.Date(max_date),
    year = as.integer(year),
    sex = tolower(sex))
```

## Criteria
```{r}
min_detection_months <- 6
min_coa_positions <- 50
```

## Apply criteria
```{r}
complete_df <- complete_df %>%
  mutate(

    # Common filtering criteria
    pass_months =
      n_detection_months >= min_detection_months,

    pass_coa_positions =
      n_kde_positions >= min_coa_positions,

    # Validity of spatial estimates
    pass_kde_50 =
      !is.na(kde_area_50) &
      kde_area_50 > 0,

    pass_kde_95 =
      !is.na(kde_area_95) &
      kde_area_95 > 0,

    pass_mcp100 =
      !is.na(mcp_area_100) &
      mcp_area_100 > 0,

    # Common eligibility for all spatial analyses
    use_spatial_analysis =
      pass_months &
      pass_coa_positions &
      pass_kde_50 &
      pass_kde_95 &
      pass_mcp100,

    # Metric-specific usable samples
    use_kde =
      use_spatial_analysis &
      pass_kde_50 &
      pass_kde_95,

    use_mcp100 =
      use_spatial_analysis &
      pass_mcp100)
```

## Checks
```{r}
filter_audit <- tibble(
  stage = c(
    "All shark-years",
    ">=6 detection months",
    ">=6 months + >=50 COAs",
    "Valid KDE sample",
    "Valid MCP100 sample",
    "Usable for at least one spatial metric"),

  n_shark_years = c(
    nrow(complete_df),

    sum(complete_df$pass_months),

    sum(complete_df$use_spatial_analysis),

    sum(complete_df$use_kde),

    sum(complete_df$use_mcp100),

    sum(
      complete_df$use_spatial_analysis &
      (complete_df$use_kde | complete_df$use_mcp100)
    )),

  n_sharks = c(
    n_distinct(complete_df$shark_ID),

    n_distinct(
      complete_df$shark_ID[
        complete_df$pass_months
      ]),

    n_distinct(
      complete_df$shark_ID[
        complete_df$use_spatial_analysis
      ]),

    n_distinct(
      complete_df$shark_ID[
        complete_df$use_kde
      ]),

    n_distinct(
      complete_df$shark_ID[
        complete_df$use_mcp100
      ]),

    n_distinct(
      complete_df$shark_ID[
        complete_df$use_spatial_analysis &
        (complete_df$use_kde | complete_df$use_mcp100)
      ]
    )
  )
)

filter_audit
```

# Export 
## Dataset for modelling and figures
```{r}
annual_spatial_df <- complete_df %>%
  filter(use_spatial_analysis) %>%
  dplyr::select(
    shark_ID,
    year,

    size,
    sex,
    maturity,

    n_detection_months,
    n_tracking_days,
    n_kde_positions,
    num_detections,
    num_receivers,

    kde_area_50,
    kde_area_95,
    mcp_area_100)

stopifnot(
  all(annual_spatial_df$n_detection_months >= 6),
  all(annual_spatial_df$n_kde_positions >= 50))

saveRDS(
  annual_spatial_df,
  "../data/intermediate/annual_space_use_6months_50coa.rds")
```

```{r}
stopifnot(
  all(annual_spatial_df$n_detection_months >= 6),
  all(annual_spatial_df$n_kde_positions >= 50)
)
```

## Datasets for maps and polygons

Retains all COA positions belonging to shark-years included in the final
annual spatial-analysis dataset.
```{r}
# COA dataset - used for inspection of filters
raw_df <- read.csv("../data/intermediate/combined_sharks_coa.csv")

coa_analysis <- raw_df %>%
  mutate(
    date = as.Date(ISO.date),
    year = lubridate::year(date)
  ) %>%
  inner_join(
    complete_df %>%
      filter(use_spatial_analysis) %>%
      dplyr::select(
        shark_ID,
        year,
        sex,
        size,
        maturity,
        kde_area_50,
        kde_area_95,
        mcp_area_100
      ),
    by = c(
      "FishID" = "shark_ID",
      "year"
    )
  )

saveRDS(
  coa_analysis,
  "../data/intermediate/coa_spatial_analysis.rds")
```
