---
title: "2023_01_25_plotting_MC"
author: "Maxine Chapman"
date: "2023-01-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Plots based on the annual results dataframe (not raw data directly).

# Setup

```{r}
rm(list =ls()) #clear environments
```


## packages

```{r}
library(ggplot2)
library(ggsignif)
library(lme4)
library(performance)
library(cowplot)
library(dplyr)
library(ggsci)
library(geomtextpath)
library(forcats)
library(r2r)
library(magick)
library(scales)
library(rnaturalearth)
library(raster)
library(emmeans)
library(ggspatial)
library(igraph)
library(ggnewscale)
```

## source functions

```{r}
source("03_data_processing_functions_MC.R") #load code to plot mixed model diagnostic plots
```

## read in data

```{r}
#read in results df
results_df <- read.csv(
  "../results/results_annual_kde_coa.csv") #full results no filtering

plot_df <- readRDS(
  "../data/intermediate/annual_space_use_6months_50coa.rds") #filtered by KUD 6 months

seasonal_df <- readRDS(
  "../results/seasonality/results_monthly_residency.rds") #seasonality df
```

```{r}
#read in receiver data
receiver_data = read.csv("../data/raw/acoustic_metadata/darros_receiver_details.csv")
#round lat and lon to 3 sf for matching with the shark data
receiver_data$Lat<-round(receiver_data$Lat,3)
receiver_data$Long<-round(receiver_data$Long,3)

#rename receiver ID column for clarity
colnames(receiver_data)[which(colnames(receiver_data) == "DSJ.Code")] <- "ReceiverID"

#head(receiver_data)
```


# Main Figures

```{r}
#make filepath
filepath2 <-file.path("..", "results", "figures")
```

## fig. 1: Receiver maps

```{r}
# Wide-area bathymetry
g2_raster <- raster::raster(
  "../resources/ETOPO_2022_bathy_extract.tiff"
)

#get coordinates for rasters
g2_extent <- c(
  xmin(g2_raster),
  xmax(g2_raster),
  ymin(g2_raster),
  ymax(g2_raster)
)

#island
map_small <- terra::rast("../resources/Georeferenced island imagery/satellite/islands11.tif")

#get coordinates
g3_extent <- c(xmin(map_small),
               xmax(map_small),
               ymin(map_small),
               ymax(map_small))
```

```{r}
#create largest map showing position in Indian Ocean
world <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf"
)

g1 <- ggplot() +
  
  # World map
  geom_sf(
    data = world,
    fill = "grey85",
    colour = "grey50",
    linewidth = 0.2
  ) +
  
  # Box showing the full extent of g2
  geom_rect(
    aes(
      xmin = g2_extent[1],
      xmax = g2_extent[2],
      ymin = g2_extent[3],
      ymax = g2_extent[4]
    ),
    fill = NA,
    colour = "red",
    linewidth = 0.8
  ) +
  
  # Label for g2 box
  annotate(
    "text",
    x = g2_extent[1],
    y = g2_extent[4],
    label = "Amirantes Bank",
    colour = "red",
    fontface = "bold",
    size = 4.5,
    hjust = 0.4,
    vjust = -0.5
  ) +
  
  # Africa label
  annotate(
    "text",
     x = 35,
    y = -5,
    label = "Africa",
    colour = "grey40",
    fontface = "bold",
    size = 4
  ) +
  
  # Indian Ocean label
  annotate(
    "text",
    x = 70,
  y = -15,
    label = "Indian Ocean",
    colour = "grey40",
    fontface = "italic",
    size = 4
  ) +
  #surroundings limits
  coord_sf(
    xlim = c(20, 85),
    ylim = c(-30, 20),
    expand = FALSE
  ) +
  theme_void() +
  theme(
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.7
    )
  ) +
  annotation_scale(text_cex = 1)

g1
```

```{r}
#amirantes bank map
g2_raster2 <- terra::rast(g2_raster)

g2 <- ggplot() +
  tidyterra::geom_spatraster(aes(fill = ETOPO_2022_bathy_extract), data = g2_raster2) +
  scale_fill_gradientn(colours = c("#152E47", "#51AEF6"), guide = "none") +
  ggnewscale::new_scale_fill() +
  geom_point(data = receiver_data, aes(x = Long, y = Lat, fill = Habitat),
             shape = 21, size = 2.5, colour = "black", stroke = 0.3) +
  geom_rect(aes(xmin = g3_extent[1], xmax = g3_extent[2], ymin = g3_extent[3], ymax = g3_extent[4]),
            fill = NA, colour = "red", linewidth = 0.8) +
  annotate("text", x = g3_extent[1], y = g3_extent[4], label = "D'Arros and\n St Joseph",
           colour = "red", fontface = "bold", size = 4.5, hjust = -0.5, vjust = 0.5) +
  annotation_scale(location = "bl", width_hint = 0.25, text_cex = 1) +
  scale_fill_manual(values = c("Lagoon" = "green", "Coastal reef" = "pink",
                               "Plateau" = "orange", "Dropoff" = "purple"),
                    name = "Habitat Type") +
  coord_sf(xlim = g2_extent[1:2], ylim = g2_extent[3:4], expand = FALSE) +
  theme_void() +
  theme(legend.position = "none")
```


```{r}
#D'Arros and St Joseph map
g3 <- ggplot() +
  tidyterra::geom_spatraster_rgb(data = map_small) +
  geom_point(data = receiver_data, aes(x = Long, y = Lat, fill = Habitat),
             shape = 21, size = 5, stroke = 0.5) +
  annotate("text", x = 53.295, y = -5.405, label = "D'Arros",
           colour = "red", fontface = "bold", size = 4.5, hjust = 0, vjust = 0) +
  annotate("text", x = 53.365, y = g3_extent[3] + 0.75 * diff(g3_extent[3:4]),
         label = "St Joseph", colour = "red", fontface = "bold",
         size = 4.5, hjust = 1, vjust = 0) +
  annotate("rect", xmin = g3_extent[1], xmax = g3_extent[1] + 0.025,
         ymin = g3_extent[3], ymax = g3_extent[3] + 0.004,
         fill = "white", colour = NA, alpha = 0.9) +
  annotation_scale(location = "bl", width_hint = 0.25, text_cex = 1) +
  scale_fill_manual(values = c("Lagoon" = "green", "Coastal reef" = "pink",
                               "Plateau" = "orange", "Dropoff" = "purple"),
                    name = "Habitat Type") +
  coord_sf(xlim = g3_extent[1:2], ylim = g3_extent[3:4], expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom", legend.text = element_text(size = 12))
```

```{r}
filename <- "figure_1_receiver_maps.png"

col1 <- cowplot::plot_grid(
  g1, g2,
  ncol = 1,
  labels = c("a", "b"),
  label_size = 14,
  rel_heights = c(0.5, 1),
  rel_widths = c(1, 1), # adjust proportions if needed
  align = "none"
)

final <- cowplot::plot_grid(
  col1, g3,
  ncol = 2,
  labels = c("", "c"),
  label_size = 14,
  rel_widths = c(1, 2),
  align = "none"
)

final
ggsave(filename, plot = final, path = filepath2, width = 10, height = 8, units = "in", dpi = 300)
rm(filename)
```

## fig. 2: Descriptive Plot

```{r}
#limit unil 2022
filt_df <- results_df %>% filter(year < 2022)

#make new columns
included_sharks <- unique(plot_df$shark_ID)

describe_df <- filt_df %>% #includes all sharks
  dplyr::mutate(log_detections = log10(num_detections)) %>%
  dplyr::group_by(shark_ID) %>%
  dplyr::mutate(start_year = min(year)) %>%
  dplyr::mutate(end_year = max(year)) %>%
  dplyr::mutate(total_days = sum(n_tracking_days)) %>%
  dplyr::mutate(included = ifelse(shark_ID %in% included_sharks, "included", "not included")) %>% 
  arrange(desc(included), shark_ID)
  #dplyr::mutate(shark_ID = factor(shark_ID, levels = unique(shark_ID)))

describe_df$shark_ID <- factor(describe_df$shark_ID, levels = unique(describe_df$shark_ID))

maturity_df <- describe_df %>%
  filter(maturity == "mature") %>%
  group_by(shark_ID) %>%
  summarize(maturity_year = min(year), .groups = "drop")

# Step 3: Merge maturity year back into main dataframe
describe_df <- left_join(describe_df, maturity_df, by = "shark_ID")


shade_df <- describe_df %>%
  filter(included == "not included") %>%
  distinct(shark_ID) %>%
  mutate(
    xmin = as.numeric(shark_ID) - 0.5,
    xmax = as.numeric(shark_ID) + 0.5,
    ymin = -Inf,   # cover all years
    ymax = Inf
  )

my_breaks <- c(1, 10, 100, 1000, 10000, 100000)
my_breaks_log <- log10(my_breaks)
```

NB: The plot appears differently (less cramped) in saved files than in R due to different dimensions in save.

```{r}
plot <- ggplot(describe_df, aes(x = factor(shark_ID), y = year, ymin = year, ymax = year + 1, fill = log_detections)) +
  geom_crossbar(aes(size = NA), width = 0.6) +  # narrower bars
  geom_text(label = "Detection days", y = 2024, x = length(unique(describe_df$shark_ID)) + 1, hjust = 0.6) + #label tracking days column
  coord_flip() +
  scale_fill_gradient(low = "#999999", high = "#377EB8", name = "Number of detections", breaks = my_breaks_log,     
  labels = comma(my_breaks)) +
  scale_y_continuous(breaks = seq(2012, 2022, by = 1), expand = expansion(mult = c(0, 0.15))) +
  #scale_y_discrete(expand = expansion(add = c(0, 2))) +
  scale_x_discrete(expand = expansion(add = c(0, 2))) +  #more horizontal space for shark_ID
  #labs(title = "Shark Tracking Data Visualisation", x = "Shark ID", y = "Year", fill = "log(number of detections)") +
  labs(x = "Shark ID", y = "Year", fill = "Number of detections)") +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(size = FALSE) +
  geom_text(aes(y = 2024, label = total_days), vjust = 0.5, size = 3.3) +
  theme_bw() +
  theme(panel.grid.minor.x = element_blank()) +
  # theme(panel.grid.major.y = element_line(linewidth = 1, colour = "grey"), 
  #       axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_rect(data = shade_df,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey80", alpha = 0.5,
            inherit.aes = FALSE) + #grey out irrelevant data
  geom_segment(
  data = maturity_df,
  aes(
    x = shark_ID,
    xend = shark_ID,
    y = maturity_year - 0.05,
    yend = maturity_year + 0.05
  ),
  color = "#F8766D",
  size = 1,
  inherit.aes = FALSE
)

plot  
#saving
filename <- "figure_2_descriptive.png"
#filename <- "figure_6_descriptive_fast.png"
ggsave(filename, path = filepath2)
```

## fig. 3: residency results

```{r}
four_group_summary <- seasonal_df %>%
  group_by(month, sex, maturity_binary) %>%
  summarise(
    mean_residency = mean(residency, na.rm = TRUE),
    se_residency = sd(residency, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    group = interaction(
      sex,
      maturity_binary,
      sep = " - "
    ),
    month_factor = factor(
      month,
      levels = 1:12,
      labels = month.abb
    )
  )

four_group_plot <- ggplot(four_group_summary,aes(
    x = month_factor,
    y = mean_residency,
    colour = sex,
    linetype = maturity_binary,
    group = group
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = pmax(0, mean_residency - se_residency),
      ymax = pmin(1, mean_residency + se_residency)
    ),
    width = 0.15
  ) +
  scale_y_continuous(
    limits = c(0, 1)
  ) + 
scale_colour_manual(
    values = c(
      "f" = "#F8766D",
      "m" = "#377EB8"
    ),
    name = "Sex"
  ) +
  scale_linetype_manual(
    values = c(
      "immature" = "dotted",
      "mature" = "solid"
    ),
    name = "Maturity"
  ) +
  labs(
    x = "Month",
    y = "Mean monthly residency"
  ) +
  theme_bw()

four_group_plot

filename <- "figure_3_array_residency_all.png"
ggsave(filename, path = filepath2, width = 7, height = 5, dpi = 600)
```

## fig. 4: multipanel model results area v shark size

Models 
```{r}
# 50% KUD
mod_50_fig <- lmer(
  log_kde_50 ~ size + sex + n_detection_months + (1 | shark_ID),
  data = mod_df, REML = TRUE)

# 95% KUD
mod_95_fig <- lmer(
  log_kde_95 ~ size + sex + n_detection_months + (1 | shark_ID),
  data = mod_df, REML = TRUE)

# 100% MCP
mod_mcp_fig <- lmer(
  log_mcp_100 ~ size + sex + n_detection_months + (1 | shark_ID),
  data = mod_df, REML = TRUE)

# Custom colour scale
sex_colours <- c("m" = "#377EB8", "f" = "#F8766D")
```

Plotting Function
```{r}
make_spaceuse_panel <- function(model, data, area_col, y_label) {
  
  size_seq <- seq(
    min(data$size, na.rm = TRUE),
    max(data$size, na.rm = TRUE),
    length.out = 100
  )
  
  pred <- emmeans(
    model,
    ~ size,
    at = list(size = size_seq),
    weights = "proportional"
  ) %>%
    confint() %>%
    as.data.frame() %>%
    mutate(
      predicted_km2 = 10^emmean,
      conf.low_km2  = 10^lower.CL,
      conf.high_km2 = 10^upper.CL
    )
  
  ggplot(data, aes(x = size, y = .data[[area_col]])) +
    geom_point(
      colour = "grey45",
      alpha = 0.6,
      size = 1.6
    ) +
    geom_ribbon(
      data = pred,
      aes(
        x = size,
        ymin = conf.low_km2,
        ymax = conf.high_km2
      ),
      inherit.aes = FALSE,
      fill = "grey70",
      alpha = 0.35
    ) +
    geom_line(
      data = pred,
      aes(
        x = size,
        y = predicted_km2
      ),
      inherit.aes = FALSE,
      colour = "black",
      linewidth = 0.9
    ) +
    scale_y_log10() +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank()
    ) +
    labs(
      x = "Shark size (cm)",
      y = y_label
    )
}
```

Panels
```{r}
kde_50 <- make_spaceuse_panel(
  mod_50_fig,
  mod_df,
  "kde_area_50",
  expression("50% KUD area (km"^2*")")
)

kde_95 <- make_spaceuse_panel(
  mod_95_fig,
  mod_df,
  "kde_area_95",
  expression("95% KUD area (km"^2*")")
)

mcp_100 <- make_spaceuse_panel(
  mod_mcp_fig,
  mod_df,
  "mcp_area_100",
  expression("100% MCP area (km"^2*")")
)
```

Assemble final plot
```{r}
library(cowplot)

# remove repeated x-axis titles from top row
kde_50_top <- kde_50 + labs(x = NULL)
kde_95_top <- kde_95 + labs(x = NULL)

# Top row
top_row <- plot_grid(
  kde_50_top,
  kde_95_top,
  labels = c("a", "b"),
  ncol = 2,
  align = "hv"
)

# Bottom panel centred, same width as either top panel
bottom_row <- plot_grid(
  NULL,
  plot_grid(
    mcp_100,
    labels = "c"
  ),
  NULL,
  ncol = 3,
  rel_widths = c(0.5, 1, 0.5)
)

# Final figure
figure_4 <- plot_grid(
  top_row,
  bottom_row,
  ncol = 1,
  rel_heights = c(1, 1)
)

figure_4

filename <- "figure_4_multipanel_models.png"
ggsave(filename, path = filepath2, width = 6.85, height = 7.5, dpi = 600)
rm(filename, top_row, bottom_row, kde_50, kde_95, mcp_100)
```


# Supplementary Figures

## Example KUDs and MCPs

NB: Need to run figure 1 section first.
```{r}
#read in data
all_sharks <- readRDS("../data/intermediate/coa_spatial_analysis.rds") #read in all sharks

selected_shark <-  59

df <- all_sharks %>% dplyr::filter(FishID %in% selected_shark)
unique(df$year)

# select_years <- c("2013", "2016", "2019") #55
select_years <- c("2014", "2018", "2021") #59
# select_years <- c("2014", "2015", "2016") #144
# select_years <- c("2016", "2018", "2020") #364
```

```{r}
#convert to correct projection
sharks_tt <- tt_read_data(
  events = df,
  col_track_id = "FishID",
  col_coords = c("Long", "Lat"),
  col_date_time = "ISO.date",
  time_zone = "UTC",
  format_date_time = "%Y-%m-%d %H:%M:%OS",
  crs = 4326)

#calculate kde
sharks_kde <- sharks_tt %>%
  group_by(FishID, year) %>%
  hr_kde(levels = c(0.5, 0.95))

#calculate mcp
sharks_mcp <- sharks_tt %>%
  group_by(FishID, year) %>%
  hr_mcp(levels = c(1.00))
```

```{r}
#make plotting df
kde_plot <- sharks_kde %>%
  mutate(HR = case_when(level == 0.95 ~ "KUD95", level == 0.50 ~ "KUD50"))

mcp_plot <- sharks_mcp %>%
  mutate(HR = "MCP100") 

#combine dfs
hr_plot <- bind_rows(kde_plot, mcp_plot) %>%
  mutate(HR = factor(HR, levels = c("KUD95", "KUD50", "MCP100")),
    Year = factor(sub(".*_", "", group_id), select_years)) %>% 
  filter(!is.na(Year))
```


```{r}
#plotting
s1 <- ggplot() +
  tidyterra::geom_spatraster_rgb(data = map_small) +
  geom_sf(data = hr_plot, aes(fill = HR), alpha = 0.5, colour = "black") + 
  facet_grid(HR ~ Year) +
  coord_sf(xlim = g3_extent[1:2], ylim = g3_extent[3:4], expand = FALSE) +
  theme_void() +
  labs(fill = "Space Use\nMetric")

s1

filename <- paste0("supplementary_polygons_", selected_shark, ".png")
ggsave(filename, plot = s1, path = filepath2)

rm(filename)
```

