---
title: "modelling"
author: "Maxine Chapman"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
rm(list =ls()) # clear environment

library(lme4) # add libraries
library(boot)
library(rstatix)
library(ggpubr)
library(numDeriv)

source("plot_lmer_diag_loess.R") #load code to plot mixed model diagnostic plots
```

## Make datasets

```{r}
complete_df <- read.csv("../results/results_dataframe_annual.csv") #read in data
head(complete_df)

plot_df <- complete_df[complete_df$n_tracking_days > 40, ] #filter for number of tracking days
plot_df <- na.omit(complete_df) #remove na values for linear modelling
head(plot_df)

#plot_df <- plot_df[plot_df$shark_ID != 368, ] #remove potential outlier
```

## Set growth rates

```{r}
#select sizes produced by either slow or fast growth rate
size <- plot_df$size_fast 
size_all <- complete_df$size_fast
```


## Fixed effect models

```{r}
lm <- lm(data = plot_df, log10(mcp_area_80+1) ~ size + log(n_tracking_days) + shark_ID + sex) #make lm with size
summary(lm) #show model output
```

```{r}
par(mfrow = c(2, 2)) #display plots correctly
plot(lm) #make diagnostic plots
```

```{r}
lm2 <- lm(data = plot_df, log10(mcp_area_80+1) ~ log(n_tracking_days) + shark_ID + sex) # make lm without size
summary(lm2) #show model output
```

```{r}
par(mfrow = c(2, 2)) #display
plot(lm2) #diagnostic plots
```

```{r}
anova(lm, lm2) #compare models
```

Shows the model with size is significantly better than the model without size, indicating that size is a significant predictor of MCP. 

## Mixed effect model

```{r}
mixed_lm <- lmer(log10(mcp_area_80+1) ~ size + log(n_tracking_days) + (1|shark_ID) + sex, data = plot_df) #full model
summary(mixed_lm) #show model output
```

```{r}
mixed_lm2 <- lmer(log10(mcp_area_80+1) ~ log(n_tracking_days) + (1|shark_ID) + sex, data = plot_df) #mixed model without size
summary(mixed_lm2) #show model output
```

```{r}
mixed_lm3 <- lmer(log10(mcp_area_80+1) ~ size + log(n_tracking_days) + (1|shark_ID), data = plot_df) #mixed model without sex
summary(mixed_lm3) #show model output
```

```{r}
anova(mixed_lm, mixed_lm2, test = "Chisq") #compare models with size
```

Shows the model with size is significantly better than the model without size, indicating that size is a significant predictor of MCP. 

```{r}
anova(mixed_lm, mixed_lm3, test = "Chisq") #compare models with sex
```

Diagnostic plots
```{r}
plot_lmer_diag(mixed_lm, color = "shark_ID") #leverage and cook's distance plots
```

```{r}
plot_lmer_rand_diag(mixed_lm, random="shark_ID") #random intercepts and normal Q-Q plots
```


# Binary model

Testing if there is significant difference in the sizes of exploratory v resident sharks:

```{r}
# make model with size
binary_glm <- glm(binary_num ~ size_all + log(n_tracking_days) + shark_ID, family = binomial, data = complete_df)
summary(binary_glm) # show output
```

```{r}
#make diagnostic plots
par(mfrow = c(2, 2))
glm.diag.plots(binary_glm)
```

```{r}
# make model without size
binary_glm2 <- glm(binary_num ~  log(n_tracking_days) + shark_ID, family = binomial, data = complete_df)
summary(binary_glm2)
```

```{r}
#make diagnostic plots
par(mfrow = c(2, 2))
glm.diag.plots(binary_glm2)
```

```{r}
# compare models
anova(binary_glm, binary_glm2, test = "Chisq" )
```

