---
title: "Growth and morphometrics"
author: "Georgios Papageorgiou"
date: "2026-08-30"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

# Growth and morphometrics

This analysis evaluates the body length measurements and growth assumptions used to estimate annual shark size. Specifically, it

1. examines the relationship between TL and PCL in the sharks of this study;
2. derives a TL–PCL conversion for comparison with published measurements reported in TL;
3. evaluates the Brown and Gruber (1988) Negaprion brevirostris growth curve against juvenile N. acutidens growth reported by Stevens (1984);
4. converts the published 220 cm TL maturity threshold to PCL.

# Setup
```{r}
library(tidyverse)
library(lubridate)
```

```{r}
capture_df <- read.csv(
  "../data/raw/acoustic_metadata/lemon_summary_read_updated.csv")

morph_df <- capture_df %>%
  transmute(
    FishID,
    species = Species,
    sex = tolower(Sex),
    pcl = as.numeric(pcl),
    fl = as.numeric(fl),
    tl = as.numeric(tl),
    tagging_date = dmy_hm(Date.released))
```

```{r}
# check the dataset, how many pcl - tl pairs we have and max-min values
morph_df %>%
  summarise(
    n_sharks = n(),
    n_paired_TL_PCL = sum(!is.na(tl) & !is.na(pcl)),
    min_TL = min(tl, na.rm = TRUE),
    max_TL = max(tl, na.rm = TRUE),
    min_PCL = min(pcl, na.rm = TRUE),
    max_PCL = max(pcl, na.rm = TRUE))

# check for duplicates
morph_df %>%
  count(FishID) %>%
  filter(n > 1)
```

# TL–PCL relationship
Restrict the morphometric analysis to sharks for which both measurements are available.
```{r}
tl_pcl <- morph_df %>%
  filter(
    !is.na(tl),
    !is.na(pcl))
```

## Linear Model
```{r}
tl_pcl_model <- lm(
  pcl ~ tl,
  data = tl_pcl)

summary(tl_pcl_model)
confint(tl_pcl_model)
```

```{r}
tl_pcl_intercept <- coef(tl_pcl_model)[1]
tl_pcl_slope <- coef(tl_pcl_model)[2]

tl_pcl_intercept
tl_pcl_slope
```

## Plot
Supplementary figure: TL-PCL relationship
```{r}
r2 <- summary(tl_pcl_model)$r.squared
n_pairs <- nrow(tl_pcl)

fig_tl_pcl <- ggplot(tl_pcl, aes(x = tl, y = pcl)) +
  geom_point(size = 2) +
  geom_smooth(
    method = "lm",
    se = TRUE) +
  annotate(
    "text",
    x = Inf,
    y = -Inf,
    hjust = 1.1,
    vjust = -0.8,
    label = paste0(
      "PCL = ", round(tl_pcl_slope, 3),
      " TL + ", round(tl_pcl_intercept, 3),
      "\nR² = ", round(r2, 3),
      ", n = ", 43)
  ) +
  labs(
    x = "Total length (cm)",
    y = "Precaudal length (cm)"
  ) +
  theme_bw()

fig_tl_pcl

ggsave(
    "../results/growth/figure_S_TL_PCL.png",
    fig_tl_pcl,
    width = 6,
    height = 5,
    dpi = 300)
```

## Model diagnostics
```{r}
par(mfrow = c(2, 2))
plot(tl_pcl_model)
par(mfrow = c(1, 1))
```

Test for curvature
```{r}
tl_pcl_quadratic <- lm(
  pcl ~ tl + I(tl^2),
  data = tl_pcl)

anova(tl_pcl_model, tl_pcl_quadratic)

AIC(tl_pcl_model, tl_pcl_quadratic)
```

Conversion functions
```{r}
TL_to_PCL <- function(TL_cm) {
  predict(
    tl_pcl_model,
    newdata = data.frame(tl = TL_cm)
  )
}

PCL_to_TL <- function(PCL_cm) {
  (PCL_cm - tl_pcl_intercept) / tl_pcl_slope
}
```

# Candidate Brown & Gruber growth model
```{r}
Linf_brown <- 317.65
k_brown <- 0.057

fabens_size <- function(
  initial_size,
  years,
  Linf = Linf_brown,
  k = k_brown
) {
  
  Linf - (Linf - initial_size) * exp(-k * years)
}
```

# Validate against Stevens (1984)

Stevens' short-term growth dataset contained 16 immature N. acutidens measuring 74.5–121.2 cm TL at release. Their mean growth rate was approximately 10.4 mm/month, or 12.5 cm/year TL.

Two longer-term recaptures were:
- 127 cm TL at tagging, growing approximately 15.4 cm/year
- 136 cm TL at tagging, growing approximately 15.6 cm/year

Because Stevens reports TL whereas Brown and Gruber parameterised growth using PCL, the local TL–PCL relationship is used to put both sources on a common scale.

```{r}
# Stevens long-term recaptures
stevens_longterm <- tibble(
  start_TL = c(127, 136),
  observed_growth_TL = c(15.4, 15.6)
) %>%
  mutate(
    start_PCL = as.numeric(TL_to_PCL(start_TL)),
    
    end_PCL = fabens_size(
      initial_size = start_PCL,
      years = 1
    ),
    
    end_TL = PCL_to_TL(end_PCL),
    
    predicted_growth_TL = end_TL - start_TL,
    
    difference = predicted_growth_TL - observed_growth_TL
  )

stevens_longterm
```

# Convert maturity threshold from TL to PCL

Stevens reported sexual maturity at approximately 220 cm TL. 
Use the locally fitted morphometric relationship to estimate the corresponding PCL:
```{r}
maturity_TL <- 220

maturity_PCL_prediction <- predict(
  tl_pcl_model,
  newdata = data.frame(
    tl = maturity_TL
  ),
  interval = "confidence")

maturity_PCL_prediction

maturity_PCL <- as.numeric(
  maturity_PCL_prediction[1, "fit"])

maturity_PCL
```