---
title: "Annual space use modelling"
author: "Giorgos Papageorgiou"
date: "2026-09-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
```{r}
rm(list =ls()) # clear environment
library(lme4)
library(tidyverse)
library(writexl)

source("plot_lmer_diag_loess.R") #load code to plot mixed model diagnostic plots
```

# Datasets
```{r}
mod_df <- readRDS(
  "../data/intermediate/annual_space_use_6months_50coa.rds")

coa_df <- readRDS(
  "../data/intermediate/coa_spatial_analysis.rds")

tables_dir <- "../results/tables"

dir.create(
  tables_dir,
  recursive = TRUE,
  showWarnings = FALSE)
```

```{r}
mod_df <- mod_df %>%
  mutate(
    shark_ID = factor(shark_ID),
    sex = factor(tolower(as.character(sex))),
    life_stage = factor(maturity),
    log_kud_50 = log10(kde_area_50),
    log_kud_95 = log10(kde_area_95),
    log_mcp_100 = log10(mcp_area_100))
```

# KUD models

Full models
```{r}
# 50% KUD - core area
mod_50 <- lmer(
  log_kud_50 ~ size + sex + n_detection_months + (1 | shark_ID),
  data = mod_df, REML = FALSE)

summary(mod_50)

# 95% KUD - home range
mod_95 <- lmer(
  log_kud_95 ~ size + sex + n_detection_months + (1 | shark_ID),
  data = mod_df, REML = FALSE)

summary(mod_95)
```

anova() tests
```{r}
# 50% KUD
mod_50_no_size <- update(mod_50, . ~ . - size)
mod_50_no_sex  <- update(mod_50, . ~ . - sex)
mod_50_no_months <- update(mod_50, . ~ . - n_detection_months)

anova(mod_50, mod_50_no_size)
anova(mod_50, mod_50_no_sex)
anova(mod_50, mod_50_no_months)

# 95% KUD
mod_95_no_size <- update(mod_95, . ~ . - size)
mod_95_no_sex  <- update(mod_95, . ~ . - sex)
mod_95_no_months <- update(mod_95, . ~ . - n_detection_months)

anova(mod_95, mod_95_no_size)
anova(mod_95, mod_95_no_sex)
anova(mod_95, mod_95_no_months)
```

Diagnostic plots
```{r}
plot_lmer_diag(mod_50) #leverage and cook's distance plots
plot_lmer_diag(mod_95)
```

```{r}
plot_lmer_rand_diag(mod_50, random="shark_ID") #random intercepts and normal Q-Q plots
```

# MCP models
```{r}
mod_mcp <- lmer(
  log_mcp_100 ~ size + sex + n_detection_months + (1 | shark_ID),
  data = mod_df,
  REML = FALSE)

summary(mod_mcp)

# Dropping each predictor
mod_mcp_no_size <- update(mod_mcp, . ~ . - size)
mod_mcp_no_sex <- update(mod_mcp, . ~ . - sex)
mod_mcp_no_months <- update(mod_mcp, . ~ . - n_detection_months)

anova(mod_mcp, mod_mcp_no_size)
anova(mod_mcp, mod_mcp_no_sex)
anova(mod_mcp, mod_mcp_no_months)
```

Diagnostics
```{r}
par(mfrow = c(1, 2))

plot(
  fitted(mod_mcp),
  resid(mod_mcp),
  xlab = "Fitted values",
  ylab = "Residuals"
)
abline(h = 0, lty = 2)

qqnorm(resid(mod_mcp))
qqline(resid(mod_mcp))

par(mfrow = c(1, 1))

isSingular(mod_mcp)
```

# Export model results table
```{r}
make_results_table <- function(model_ml, metric) {
  
  # Refit full model with REML for coefficient estimates
  model_reml <- update(model_ml, REML = TRUE)
  
  # Fixed-effect estimates and SE
  sm <- coef(summary(model_reml))
  
  # Profile 95% confidence intervals
  ci <- confint(
    model_reml,
    parm = rownames(sm),
    method = "profile")
  
  # Likelihood-ratio test helper
  get_lrt <- function(term) {
    
    reduced <- update(
      model_ml,
      formula. = as.formula(paste(". ~ . -", term)))
    
    a <- anova(reduced, model_ml)
    
    c(
      chisq = a$Chisq[2],
      df    = a$Df[2],
      p = a$`Pr(>Chisq)`[2])
  }
  
  # LRTs from ML models
  lrt_size   <- get_lrt("size")
  lrt_sex    <- get_lrt("sex")
  lrt_months <- get_lrt("n_detection_months")
  
  # Fixed-effect names
  sex_coef <- grep("^sex", rownames(sm), value = TRUE)[1]
  
  coef_names <- c(
    "(Intercept)",
    "size",
    sex_coef,
    "n_detection_months")
  
  labels <- c(
    "Intercept",
    "Size",
    "Sex (male)",
    "Detection months")
  
  chisq <- c(NA, lrt_size["chisq"], lrt_sex["chisq"], lrt_months["chisq"])
  df <- c(NA, lrt_size["df"], lrt_sex["df"], lrt_months["df"])
  p <- c(NA, lrt_size["p"], lrt_sex["p"], lrt_months["p"])
  
  data.frame(
    Metric = c(metric, "", "", ""),
    Parameter = labels,
    
    Estimate = sprintf("%.4f", sm[coef_names, "Estimate"]),
    SE = sprintf("%.4f", sm[coef_names, "Std. Error"]),
    
    `95% CI` = paste0("(", sprintf("%.4f", ci[coef_names, 1]),", ",sprintf("%.4f", ci[coef_names, 2]),")"),
    
    `χ²` = ifelse(
      is.na(chisq),
      "-",
      sprintf("%.2f", chisq)),
    
    df = ifelse(
      is.na(df),
      "-",
      sprintf("%.0f", df)),
    
    P = ifelse(
      is.na(p),
      "-",
      ifelse(
        p < 0.001,
        "<0.001",
        sprintf("%.3f", p))),
    
    check.names = FALSE)
}

# Generate all three sections
table_50 <- make_results_table(mod_50, "50% KUD")
table_95 <- make_results_table(mod_95, "95% KUD")
table_mcp <- make_results_table(mod_mcp, "100% MCP")

# Combine
model_results_table <- rbind(
  table_50,
  table_95,
  table_mcp)

# Save to Excel
writexl::write_xlsx(
  model_results_table, file.path(
    tables_dir,
  "model_results_table.xlsx"))
```

# Piecewise analysis

Fit models
95%
```{r}
maturity_size <- 166

mod_df <- mod_df %>%
  mutate(
    size_maturity = size - maturity_size,
    pre_maturity  = pmin(0, size_maturity),
    post_maturity = pmax(0, size_maturity))

mod_95_maturity_base <- lmer(
  log_kud_95 ~ size_maturity + sex + n_detection_months +
    (1 | shark_ID),
  data = mod_df,
  REML = FALSE)

mod_95_broken_stick <- lmer(
  log_kud_95 ~ pre_maturity + post_maturity +
    sex + n_detection_months +
    (1 | shark_ID),
  data = mod_df,
  REML = FALSE)

anova(mod_95_maturity_base, mod_95_broken_stick)
summary(mod_95_broken_stick)
```

50%
```{r}
mod_50_maturity_base <- lmer(
  log_kud_50 ~ size_maturity + sex + n_detection_months +
    (1 | shark_ID),
  data = mod_df,
  REML = FALSE)

mod_50_broken_stick <- lmer(
  log_kud_50 ~ pre_maturity + post_maturity +
    sex + n_detection_months +
    (1 | shark_ID),
  data = mod_df,
  REML = FALSE)

anova(mod_50_maturity_base, mod_50_broken_stick)
summary(mod_50_broken_stick)
```

MCP
```{r}
mcp_maturity_base <- lmer(
  log_mcp_100 ~ size_maturity + sex + n_detection_months +
    (1 | shark_ID),
  data = mod_df,
  REML = FALSE)

mcp_broken_stick <- lmer(
  log_mcp_100 ~ pre_maturity + post_maturity +
    sex + n_detection_months +
    (1 | shark_ID),
  data = mod_df,
  REML = FALSE)

anova(mcp_maturity_base, mcp_broken_stick)
summary(mcp_broken_stick)
```

Piecewise model table
Function: reports two slopes and associated CIs: below and above maturity threshold
```{r}
make_broken_stick_table <- function(base_model_ml,
                                    broken_model_ml,
                                    metric) {
  # Estimates, SE and CI:
  # refit the full broken-stick model using REML
  model_reml <- update(broken_model_ml, REML = TRUE)

  sm <- coef(summary(model_reml))

  ci <- confint(
    model_reml,
    parm = rownames(sm),
    method = "profile")

  # Helper: likelihood-ratio test for individual terms
  # Uses ML model (REML = FALSE)

  get_lrt <- function(term) {

    reduced <- update(
      broken_model_ml,
      formula. = as.formula(
        paste(". ~ . -", term)))

    a <- anova(reduced, broken_model_ml)

    c(
      chisq = a$Chisq[2],
      df = a$Df[2],
      p = a$`Pr(>Chisq)`[2])
  }
  
  # Individual fixed-effect tests
  lrt_below  <- get_lrt("pre_maturity")
  lrt_above  <- get_lrt("post_maturity")
  lrt_sex    <- get_lrt("sex")
  lrt_months <- get_lrt("n_detection_months")

  # Main maturity-threshold test:
  # one slope vs broken stick
  a_change <- anova(
    base_model_ml,
    broken_model_ml)

  change_chisq <- a_change$Chisq[2]
  change_df    <- a_change$Df[2]
  change_p     <- a_change$`Pr(>Chisq)`[2]

  # Coefficient names
  sex_coef <- grep(
    "^sex",
    rownames(sm),
    value = TRUE)[1]

  coef_names <- c(
    "(Intercept)",
    "pre_maturity",
    "post_maturity",
    sex_coef,
    "n_detection_months")

  labels <- c(
    "Intercept (166 cm)",
    "Slope below maturity",
    "Slope above maturity",
    "Sex (male)",
    "Detection months")

  chisq <- c(NA, lrt_below["chisq"], lrt_above["chisq"], lrt_sex["chisq"], lrt_months["chisq"])
  df <- c(NA, lrt_below["df"], lrt_above["df"], lrt_sex["df"], lrt_months["df"])
  p <- c(NA, lrt_below["p"], lrt_above["p"], lrt_sex["p"], lrt_months["p"])

  # Full model coefficient rows
  out <- data.frame(
    Metric = c(metric, "", "", "", ""),
    Parameter = labels,

    Estimate = sprintf(
      "%.4f",
      sm[coef_names, "Estimate"]),

    SE = sprintf(
      "%.4f",
      sm[coef_names, "Std. Error"]),

    `95% CI` = paste0(
      "(",
      sprintf("%.4f", ci[coef_names, 1]),
      ", ",
      sprintf("%.4f", ci[coef_names, 2]),
      ")"),

    `χ²` = ifelse(
      is.na(chisq),
      "-",
      sprintf("%.2f", chisq)),
    
    df = ifelse(
      is.na(df),
      "-",
      sprintf("%.0f", df)),

    P = ifelse(
      is.na(p),
      "-",
      ifelse(
        p < 0.001,
        "<0.001",
        sprintf("%.3f", p))),

    check.names = FALSE)

  # Extra row: does slope change at 166 cm?
  change_row <- data.frame(
    Metric = "",
    Parameter = "Change in slope at 166 cm",
    Estimate = "-",
    SE = "-",
    `95% CI` = "-",

    `χ²` = sprintf(
      "%.2f",
      change_chisq),

    df = sprintf(
      "%.0f",
      change_df),

    P = ifelse(
      change_p < 0.001,
      "<0.001",
      sprintf("%.3f", change_p)),

    check.names = FALSE)
  
  rbind(out, change_row)
}
```

Run function
```{r}
table_bs_50 <- make_broken_stick_table(
  mod_50_maturity_base,
  mod_50_broken_stick,
  "50% KUD")

table_bs_95 <- make_broken_stick_table(
  mod_95_maturity_base,
  mod_95_broken_stick,
  "95% KUD")

table_bs_mcp <- make_broken_stick_table(
  mcp_maturity_base,
  mcp_broken_stick,
  "100% MCP")

broken_stick_results_table <- rbind(
  table_bs_50,
  table_bs_95,
  table_bs_mcp)
```

```{r}
writexl::write_xlsx(
  broken_stick_results_table,
  file.path(
    tables_dir,
    "broken_stick_model_results_table.xlsx"))
```

Results Report
```{r}
adult_effect_10cm <- function(model_ml, metric) {

  # Refit with REML so estimates/CIs match the supplementary table
  model_reml <- update(model_ml, REML = TRUE)

  # Adult slope (above 166 cm)
  beta <- fixef(model_reml)["post_maturity"]

  # Profile 95% CI for adult slope
  ci <- confint(
    model_reml,
    parm = "post_maturity",
    method = "profile"
  )

  # Convert log10 slope to % change in area per 10 cm
  pct <- (10^(10 * beta) - 1) * 100

  pct_low  <- (10^(10 * ci[1]) - 1) * 100
  pct_high <- (10^(10 * ci[2]) - 1) * 100

  data.frame(
    Metric = metric,
    Beta_adult = beta,
    Percent_per_10cm = pct,
    CI_low = pct_low,
    CI_high = pct_high
  )
}
```

```{r}
adult_50 <- adult_effect_10cm(
  mod_50_broken_stick,
  "50% KUD")

adult_95 <- adult_effect_10cm(
  mod_95_broken_stick,
  "95% KUD")

adult_mcp <- adult_effect_10cm(
  mcp_broken_stick,
  "100% MCP")

adult_effects <- rbind(
  adult_50,
  adult_95,
  adult_mcp)
```

# Sensitivity analysis: influential observations

Observation 85 had the highest Cook's distance in the 50% and 95% KUD model.
Refit without this observation to assess whether it influenced inference.
```{r}
row85 <- rownames(model.frame(mod_95))[85]
mod_df_no85 <- mod_df[rownames(mod_df) != row85, ]

sensitivity_lrt <- function(response, metric) {
  
  full <- lmer(
    as.formula(paste(
      response,
      "~ size + sex + n_detection_months + (1 | shark_ID)"
    )),
    data = mod_df_no85,
    REML = FALSE)
  
  bind_rows(lapply(
    c("size", "sex", "n_detection_months"),
    function(term) {
      reduced <- update(
        full,
        formula. = as.formula(paste(". ~ . -", term)))
      
      a <- anova(reduced, full)
      
      tibble(
        metric = metric,
        predictor = term,
        chisq = a$Chisq[2],
        p = a$`Pr(>Chisq)`[2])}
  ))
}

sensitivity_results <- bind_rows(
  sensitivity_lrt("log_kud_50", "50% KDE"),
  sensitivity_lrt("log_kud_95", "95% KDE"),
  sensitivity_lrt("log_mcp_100", "100% MCP"))

sensitivity_results
```
