---
title: "Monthly residency and seasonality"
author: "Maxine Chapman"
date: "2026-08-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
```{r}
rm(list =ls())
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(ggsci)
library(stringr)
library(glmmTMB)
library(DHARMa)
library(marginaleffects)
```

```{r}
df <- read.csv("../data/intermediate/combined_sharks.csv") # read in all sharks, raw data, not COA
df <- df %>%
  mutate(
    datetime = lubridate::parse_date_time(
      ISO.date,
      orders = c("ymd HMS", "ymd HM", "ymd"),
      tz = "UTC"),
    date = as.Date(datetime),
    year = lubridate::year(datetime),
    month = lubridate::month(datetime))

stopifnot(!any(is.na(df$datetime)))
```

Atoll bounds
```{r}
# Atoll receiver bounds
atoll_lon_min <- 53.312212
atoll_lon_max <- 53.38
atoll_lat_min <- -5.46
atoll_lat_max <- -5.39

# Restrict detections to receivers within the atoll
residency_df <- df %>%
  filter(
    Long > atoll_lon_min,
    Long < atoll_lon_max,
    Lat > atoll_lat_min,
    Lat < atoll_lat_max)
```

Metadata
```{r}
#add in shark metadata
summary_sharks <- read.csv("../data/raw/acoustic_metadata/lemon_summary_read_updated.csv")

# standardise sex labels
summary_sharks <- summary_sharks %>%
  mutate(Sex = str_trim(Sex) %>% str_to_lower(),
         Sex = factor(Sex, levels = c("f", "m")))

# shark-level tracking metadata
shark_metadata <- readRDS(
  "../results/metadata/shark_metadata_full.rds")
```

```{r}
source("03_data_processing_functions_MC.R")
```

# Monthly atoll residency

## Calculate monthly residency
```{r}
data_names_list <- unique(df$FishID) # Sharks included in the detection dataset
years_list <- c(2012:2021) # study years
months_list <- 1:12 # Calendar months

# Standardise metadata variables
shark_metadata <- shark_metadata %>%
  mutate(
    tagging_date = as.Date(tagging_date),
    final_date   = as.Date(final_date),
    sex = str_to_lower(str_trim(sex)))

results_list <- list()
row_counter <- 1

for (i_shark in seq_along(data_names_list)) { # Loop through each tagged shark

  shark_ID <- data_names_list[i_shark] #get shark data
  shark_file <- residency_df %>% # Atoll detections for this shark
    filter(FishID == shark_ID)
  
  shark_meta <- shark_metadata %>% # Retrieve this shark's metadata
    filter(shark_ID == !!shark_ID)
  
  if (nrow(shark_meta) != 1) {
    stop("Metadata problem for shark ", shark_ID)} # Each shark should have exactly one row in the metadata table
  
  # Observed tracking period
  tagging_date <- shark_meta$tagging_date 
  final_date <- shark_meta$final_date

  # Sex
  shark_sex <- shark_meta$sex
  
  # Restrict the overall study years to years during which
  # this shark was within its observed tracking period
  shark_years <- years_list[
    years_list >= lubridate::year(tagging_date) &
    years_list <= lubridate::year(final_date)]
  
  for(i_year in shark_years) { # Loop through relevant years for this shark

    # calculate shark size using function
    shark_size <- calc_shark_size(shark_ID, i_year, summary_sharks)
    
    # Life stage based on estimated PCL
    maturity <- case_when(
      shark_size >= 166   ~ "mature",
      TRUE                ~ "immature")
    
    for(i_month in months_list){ # Loop through all 12 calendar months
      
      month_start <- as.Date(
        sprintf("%04d-%02d-01",i_year,i_month))
      
      month_end <- ceiling_date(
        month_start, unit = "month") - days(1)

      month_length <- days_in_month(month_start)
      
      # Skip months outside the observed tracking period
      if (month_end < tagging_date ||
          month_start > final_date) {
        next}
      
      # Use only complete months within the observed tracking period
      if (month_start <= tagging_date || month_end >= final_date) {
        next}
      
      subset_data <- shark_file %>% 
        filter(date >= month_start, date <= month_end)

      n_tracking_days <- n_distinct(subset_data$date)
      monthly_residency <- n_tracking_days / month_length
      
      # Save this shark-month
      results_list[[row_counter]] <- data.frame(

        shark_ID = shark_ID,
        year = i_year,
        month = i_month,

        month_start = month_start,
        month_end = month_end,
        n_tracking_days = n_tracking_days,
        month_length = month_length,
        residency = monthly_residency,
        size = shark_size,
        maturity = maturity,
        sex = shark_sex)
      
      # Move to next output row
      row_counter <- row_counter + 1
    }
  }
}
# Combine all shark-month records into one dataframe
results_df <- bind_rows(results_list)
```

# Seasonality models
```{r}
seasonal_df <- results_df %>%
  mutate(
    n_not_detected = month_length - n_tracking_days, # days without detections
    month_angle = 2 * pi * (month - 1) / 12, # Cyclic representation of month
    sin_month = sin(month_angle),
    cos_month = cos(month_angle),
    sex = factor(sex, levels = c("f", "m")),     # Standardise sex
    maturity = factor(maturity, levels = c("immature", "mature")))
```

Save monthly residency data
```{r}
seasonality_dir <- file.path("..", "results/seasonality")
figures_dir <- file.path("..", "results/figures")

dir.create(seasonality_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

saveRDS(
  seasonal_df,
  file.path(seasonality_dir, "results_monthly_residency.rds"))
write.csv(
  seasonal_df,
  file.path(seasonality_dir, "results_monthly_residency.csv"),
  row.names = FALSE)
```

## Three-way interaction model
```{r}
m_3way <- glmmTMB(
  cbind(
    n_tracking_days,
    n_not_detected) ~
    (sin_month + cos_month) * sex * maturity # Seasonality × sex × maturity
    + (1 | shark_ID), 

  family = betabinomial(link = "logit"),
  data = seasonal_df)

summary(m_3way)
```

## Two-way interaction model
```{r}
m_2way <- glmmTMB(
  cbind(
    n_tracking_days,
    n_not_detected) ~
    sin_month + cos_month + sex + maturity +
    sin_month:sex +
    cos_month:sex +
    sin_month:maturity +
    cos_month:maturity +
    sex:maturity +
    (1 | shark_ID),
  family = betabinomial(link = "logit"),
  data = seasonal_df)

anova(m_2way, m_3way) # Compare two-way and three-way interaction structures
AIC(m_2way, m_3way)
```

## Overall seasonality 
```{r}
# Reduced model with no seasonal terms
m_no_season <- glmmTMB(
  cbind(
    n_tracking_days,
    n_not_detected) ~
    sex * maturity +
    (1 | shark_ID),
  family = betabinomial(link = "logit"),
  data = seasonal_df)

# Is there evidence of seasonality overall?
lrt_season <- anova(
  m_no_season,
  m_2way)

lrt_season
```

## Seasonality x sex
Does seasonality differ between males and females?
```{r}
m_no_season_sex <- glmmTMB(
  cbind(
    n_tracking_days,
    n_not_detected) ~
    sin_month +
    cos_month +
    sex +
    maturity +
    sin_month:maturity +
    cos_month:maturity +
    sex:maturity +
    (1 | shark_ID),
  family = betabinomial(link = "logit"),
  data = seasonal_df)

lrt_season_sex <- anova(
  m_no_season_sex,
  m_2way)

lrt_season_sex
```

## Seasonality x maturity
Does seasonality differ between mature and immature sharks?
```{r}
m_no_season_maturity <- glmmTMB(
  cbind(
    n_tracking_days,
    n_not_detected) ~
    sin_month + cos_month + sex + maturity +
    sin_month:sex +
    cos_month:sex +
    sex:maturity +
    (1 | shark_ID),
  family = betabinomial(link = "logit"),
  data = seasonal_df)

lrt_season_maturity <- anova(
  m_no_season_maturity,
  m_2way)

lrt_season_maturity
```

## Sex x maturity
Does the relationship between maturity and residency differ by sex?
```{r}
m_no_sex_maturity <- glmmTMB(
  cbind(
    n_tracking_days,
    n_not_detected) ~
    sin_month + cos_month + sex + maturity +
    sin_month:sex +
    cos_month:sex +
    sin_month:maturity +
    cos_month:maturity +
    (1 | shark_ID),
  family = betabinomial(link = "logit"),
  data = seasonal_df)

lrt_sex_maturity <- anova(
  m_no_sex_maturity,
  m_2way)

lrt_sex_maturity
```

## Diagnostics
```{r}
res_final <- simulateResiduals(m_2way, n = 2000)
plot(res_final)
testDispersion(res_final)
testUniformity(res_final)
testOutliers(res_final)
```

# Descriptive summaries
```{r}
# Monthly residency
raw_months <- seasonal_df %>%
  group_by(month) %>%
  summarise(
    mean_residency = mean(residency, na.rm = TRUE),
    sd = sd(residency, na.rm = TRUE),
    n = sum(!is.na(residency)),
    se = sd / sqrt(n),
    .groups = "drop"
  ) %>%
  mutate(month_name = month.abb[month])

raw_months
```

```{r}
# Residency by sex and maturity
raw_groups <- seasonal_df %>%
  group_by(sex, maturity) %>%
  summarise(
    mean_residency = mean(residency, na.rm = TRUE),
    sd = sd(residency, na.rm = TRUE),
    n = sum(!is.na(residency)),
    se = sd / sqrt(n),
    .groups = "drop")

raw_groups
```

```{r}
# Monthly range by sex and maturity
raw_group_months <- seasonal_df %>%
  group_by(month, sex, maturity) %>%
  summarise(
    mean_residency = mean(residency, na.rm = TRUE),
    .groups = "drop")

raw_group_ranges <- raw_group_months %>%
  group_by(sex, maturity) %>%
  summarise(
    minimum_monthly_mean = min(mean_residency, na.rm = TRUE),
    maximum_monthly_mean = max(mean_residency, na.rm = TRUE),
    .groups = "drop")

raw_group_ranges
```

# Model predictions
Monthly residency
```{r}
# Prediction grid:
# each sex × maturity group evaluated at each of the 12 months
month_grid <- expand_grid(
  month = 1:12,
  sex = levels(seasonal_df$sex),
  maturity = levels(seasonal_df$maturity)) %>%
  mutate(
    sin_month = sin(2 * pi * (month - 1) / 12),
    cos_month = cos(2 * pi * (month - 1) / 12),

    # Required because shark_ID occurs in the model.
    # Random effects are excluded below with re.form = NA.
    shark_ID = seasonal_df$shark_ID[1])

group_means <- avg_predictions(
  m_2way,
  newdata = month_grid,
  by = c("sex", "maturity"),
  type = "response",
  re.form = NA)

group_means
```

Overall monthly predictions
```{r}
season_months <- avg_predictions(
  m_2way,
  newdata = month_grid,
  by = "month",
  type = "response",
  re.form = NA)

season_months %>%
  as.data.frame() %>%
  mutate(month_name = month.abb[month]) %>%
  arrange(desc(estimate)) %>%
  dplyr::select(month, month_name, estimate, conf.low, conf.high)
```

Female vs male within each maturity class
```{r}
sex_contrasts <- avg_comparisons(
  m_2way,
  variables = "sex",
  by = "maturity",
  newdata = month_grid,
  type = "response",
  re.form = NA)

sex_contrasts
```

Immature vs mature within each sex
```{r}
maturity_contrasts <- avg_comparisons(
  m_2way,
  variables = "maturity",
  by = "sex",
  newdata = month_grid,
  type = "response",
  re.form = NA)

maturity_contrasts
```